\newif\ifpdf
\ifx\pdfoutput\undefined
\pdffalse % we are not running PDFLaTeX
\else
\pdfoutput=1 % we are running PDFLaTeX
\pdftrue
\fi
\ifpdf
        \documentclass[pdftex]{wsheet}
\else
        \documentclass{wsheet}
\fi
\usepackage{rcs}
\RCS $Id: spacewar.ltx,v 1.15 2001/10/07 13:52:49 paul Exp paul $
\RCS $Date: 2001/10/07 13:52:49 $
\RCS $Revision: 1.15 $
\usepackage{graphics}
\graphicspath{{../figures/}}
\usepackage[colorlinks]{hyperref}
\sheet{Game}{Minesweeper}
\author{Rhodri James}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Rhodri James. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%-----------------------------------------------------------------------------
\section{Introduction}
This is a Python games worksheet.  When you've finished it, you'll have
written a version of a classic computer game called Minesweeper.

%-----------------------------------------------------------------------------

\section{What you need to know}
\begin{itemize}
\item The basics of Python (from Sheets 1 and 2)
\item Functions (from Sheet~3; you might want to look at Sheet~F too)
\item Classes and Objects (from Sheet~O)
\end{itemize}

\begin{note}
You should definitely look at Sheet~O before starting this worksheet,
even if you did the Beginners' course, as this sheet assumes you know
about classes and objects, which aren't in that course.
\end{note}

%-----------------------------------------------------------------------------
\section{What is Minesweeper?}

Minesweeper is a simple game played on a grid.  Your job as the player
is to find all the mines hidden in the grid without stepping on them.
You may be familiar with versions of the game supplied free with some
computers --- now you'll be able to create your own version.

When the game starts, all you see is a blank grid.  You have to pick a
square to look at; if it's a mine, then it explodes and the game is
over.  If it isn't, then you find out how many mines there are in the
surrounding squares.  From this, you can eventually work out where the
mines are, and you can mark them with flags.

Once you have flagged all the mines and cleared all the other squares,
then you win the game.

To speed things up for the player, it is possible to clear out known
squares more quickly.  If, for example, you know that there is only
one mine next to a particular square, and you already know where that
mine is, then you can clear all the other squares around the square
you know about in one go.

\section{What do we need?}

So what things are there on the screen?  From the brief description
above, we can see that we will need the following things:

\begin{itemize}
\item a grid of squares,
\item flags to mark where you think mines are,
\item mines (when they explode, at least),
\item some indication of how many mines are next to each square.
\end{itemize}

The LiveWires games library provides us with the first of these, as
well as shapes that we can put on the screen to behave as mines, flags
and numbers.  These things are provided as classes, as you might
expect from Sheet~O.  We will need to make sub-classes of these things
in order to get them to know exactly what we can do.

\section{Boxes on the Screen}

While the games library does provide a class to represent boxes on the
screen, there are a few more things we would like our boxes to do for
us.  This means we are going to have to make a sub-class to get
exactly what we want.

In fact, when you think about our squares on the screen, we actually
have two distinct types of square that behave very differently some
times.  In particular, we have squares which have mines in them, which
explode when we select them, and we have squares which don't have
mines in them, which have to count up how many of their neighbours
contain mines.

This suggests that we are going to want to make subclasses of our
subclass!  Fortunately, Python has no problem with that.

\subsection{In The Beginning}

Before we get to the interesting stuff, we should tell Python that we
are going to need the various libraries that we have mentioned.  It's
a good idea also to keep all the constants that you will want to refer
to together at the beginning.  That way, you can find them easily when
you want to change them.

Type the following into the editor window (the editor window is the
one without the |>>>| characters before each line.  In this window,
Python doesn't run what you type as soon as you type it, but lets you
type a lot of things to be run when you tell it).

\begin{program}
from livewires import games
from livewires import colour
from livewires import boards
from random import randint
\end{program}

First, we tell Python we wanted to access the things in the LiveWires
|games| and |boards| modules, and also the colour names supplied by
the |colour| module.  We will also want to use the function |randint|
from the |random| module; this is in fact what we disguised as
|random_between| in the beginners' library, so you can guess what
we're going to do with that.

\begin{program}
BOX_SIZE = 40
MARGIN = 60

MINE_SIZE = BOX_SIZE/2 - 3

NUMBER_SIZE = 24
TEXT_SIZE = 18

FLAG = ((BOX_SIZE/4, 2), (BOX_SIZE/2, BOX_SIZE-2), (3*BOX_SIZE/4, 2))
\end{program}

Then we set up |BOX_SIZE| to hold the height and width of each box on
the screen, |MARGIN| to hold the gap between the boxes and the edge of
the screen, and various other useful numbers.  We did this so that we
can easily change the number of boxes if we want to, just by changing
the program at this point.  If we decide that boxes of 40 units on
each side are too small and that we want them twice as big, all we
have to do is to change that line to read |BOX_SIZE = 80| and we can
do everything automatically.  If we had just used the number 40
everywhere we were referring to the height and width of our boxes, we
would have to track down every time it was used and change them all
individually, which would be a pain.

Worse, there may be times when our box size is used to determine
another number, and we don't spot it.  |MINE_SIZE| is a good example
of that.  We will use it as the radius of the circle that we will draw
to represent a mine; clearly that needs to be a bit less than half the
width of the square, so as to look neat.  If we were just typing magic
numbers all over the place, we would have replaced it with 17 (or
whatever we thought looked neat), and wouldn't have thought twice
about it when we changed the size of our boxes.  This would look a bit
silly.  Since we're naming our constants, however, we can calculate
the value we want from |BOX_SIZE| and not be caught out.

Another thing which we're doing here is following a convention for
naming our variables.  All of the variables that we have created here
are ones that we don't intend to change during the program.  We have
given them names that are ALL IN CAPITALS AND UNDERLINES to give us a
hint that they are constants.  Python {\em doesn't} force us to do
this, but doing it makes it easier for someone reading our program to
see what is going on.  It also gives us a hint that we shouldn't be
changing the variable after we have set it up the first time.

\subsection{Making Boxes}

Now we get on to our boxes on the screen.

\begin{program}
class GridBox(boards.GameCell):
  def __init__(self, screen, i, j):
    self.init_gamecell(screen, i, j)
    self.flagged = 0
    self.shown = 0
    self.image = None
    self.flag = None
\end{program}

What this says is that our |GridBox| is a kind of |GameCell| from the
|boards| module, and that we mostly use the same parameters to create
one.  We also set up some things that we will need later --- variables
to tell us if we have marked the box with a flag or selected it to
show to the player, and variables to hold the flag shape on the screen
itself and whatever else (a number or a mine) needs to be shown.

Notice that we've got another convention here.  We've given our class
a name that has capital letters at the start of each word and no
underlines separating them.  Once again, Python doesn't care what we
call the class, it's just a trick to make it easier for us to notice
what is going on.

\subsection{Mine, All Mine}

Before we go much further, we should work out how to display a mine on
the screen.  This is something we ought to do at the end of the game
so that the player can see where he was right and where he was wrong.
This is simply a matter of drawing the mine on the board as a suitably
sized circle.  It would make it more obvious if we turned the rest of
the box red, but that's just a detail.  More importantly, if there is
a flag in the box, we should make sure it is on top of everything else
so that it can be seen.  The following code will do the trick:

\begin{program}
class Mine(boards.Container, GridBox):
  def __init(self, board, i, j):
    self.init_gridbox(board, i, j)
    self.init_container(['image', 'flag'])

  def reveal_mine(self):
    self.set_colour(colour.red)
    self.image = games.Circle(self.screen,
                              self.xpos() + BOX_SIZE/2,
                              self.ypos() + BOX_SIZE/2,
                              MINE_SIZE,
                              colour.black,
                              filled=1)
    if self.flagged:
      self.flag.raise_object()
\end{program}

All of this assumes that we keep |flagged| and |flag| up to date, so
we'd better remember to do that!

Notice that while we've said that a |Mine| is a kind of |GridBox|,
we've also declared it to be a kind of something called a
|boards.Container|.  This is a special class that the |boards| library
provides to help when an object on the screen has things that will be
displayed on top of it.  Whenever our |Mine| is moved (which won't
happen) or moved to the front of the screen (which the library will
secretly do behind our backs), the |Container| class will make sure
that everything in the box stays on top of the |Mine| itself.  In our
case that means the circle (|image|) representing the actual mine, and
the shape (|flag|) that we'll use for a flag.

If this looks worryingly complicated, don't worry.  The whole point of
the |boards.Container| class is that it does all the work for you!

We aren't finished with the |Mine| class yet, but we'll put it aside
for now.  Instead, consider for a moment what an empty box would do if
we asked it to show its mine.  That's right, it wouldn't do anything!

\begin{trythis}
If you remember, Python's way of saying ``do nothing'' is the
instruction |pass|.  Armed with this, can you make a class called
|EmptyBox| just like |Mine| above, but handling boxes which aren't
mines?
\end{trythis}


\subsection{The Main Event}

Finally, we need our class that represents the whole board, the
collection of all the boxes and what to do with them.  This will be a
subclass of the |SingleBoard| class from the library, but it won't
work in quite as simple a way as the library expects.  Type in the
following code to start with:

\begin{program}
class Minesweeper(boards.SingleBoard):
  def __init__(self, n_cols, n_rows, mines_to_find):
    self.mines_to_find = mines_to_find
    self.flag_count = 0
    self.unknown_count = n_cols * n_rows
    self.game_over = 0

    self.init_singleboard((MARGIN, MARGIN), n_cols, n_rows, BOX_SIZE)
                        
\end{program}

This just declares |Minesweeper| to be a kind of |SingleBoard|, records
the things that we will need to keep track of, and initialises the
|SingleBoard| parts of our new class.  Once we have done that, we need
to place the mines in the grid.  This means that for each mine we need
to keep picking boxes until we find one that doesn't already have a
mine in it.  Unfortunately, Python loops always carry out their test
at the beginning of the loop when we really want to do the test at the
end here.  This leads to the following rather horrid piece of code,
which you need to type immediately after what you have typed above.

\begin{program}
\C{Remember to indent correctly}
    for m in xrange(mines_to_find):
      while 1:
        i = randint(0, n_cols - 1)
        j = randint(0, n_rows - 1)
        if self.grid[i][j] == None:
          break
      self.grid[i][j] = Mine(self, i, j)
\end{program}

What this says is ``for each mine, loop forever picking a box and if
it hasn't got anything in it break out of the loop.''  Not pretty, but
it works.

Once we have done this, we need to go through the rest of the boxes in
the grid turning them into EmptyBox objects.  Once we have done that,
there are two more useful things that we can do.  First, the library
can search through the grid for us making a list of which boxes are
the neighbours of which and putting the results in a list called
|neighbours| in each box.  Second, the library will run a pointer for
us that we call a ``cursor.''  This puts a red outline around the box
that we want to look at or put a flag on, and moves it around for us
according to the left, right, up and down arrow keys.

\begin{program}
\C{Align this with the previous ``for''}
    for i in xrange(n_cols):
      for j in xrange(n_rows):
        if self.grid[i][j] == None:
          self.grid[i][j] = EmptyBox(self, i, j)

    self.create_neighbours()
    self.enable_cursor(0, 0)
\end{program}

We still need to do a little work to get the libraries to do exactly
what we want.  When we first initialise the |SingleBoard|, it will try
to create the boxes for itself.  Since we want to create the boxes in
our own peculiar way, we have to tell it not to do the work.  We do
this by overriding the function |new_gamecell|, which should normally
returns a box of the type we want.  In this case, |None| will do just
fine, since we do the rest of the work later.

We now nearly have enough to run the program and see something
happen.  What we will do, just to see it run, is to pretend that we
exploded a mine as soon as we started and end the game, showing all
the mines.  To do this we need to write one last function in the
|Minesweeper| class and a separate useful function:

\begin{program}
\C{Align with the ``def __init__''}
  def new_gamecell(self, i, j):
    return None

  def end_game(self):
    self.game_over = 1
    self.map_grid(reveal_all_mines)
    self.disable_cursor()

def reveal_all_mines(cell):
  cell.reveal_mine()

minefield = Minesweeper(8, 8, 10)
minefield.end_game()
\end{program}

Now run it and see what happens!  The function |map_grid| that we call
in |end_game| is a useful little thing in the library that calls the
function that it is given on each box in the grid in turn.


\section{See What's There}

So far we can blow up immediately after we start.  Fun as that is,
it's not much of a game.  Let's see about searching boxes to see what
is in them.

The first thing we need to do is to decide how we want the player to
tell us to search a box.  The easiest thing to do would be for the
player to press some key when the cursor is over the box in question;
arbitrarily we pick on the Return key.

When the |boards| module sees the player press a key that it doesn't
understand (i.e. that isn't one of the arrow keys), it calls the
function |handle_keypress| to tell us, sending us a code to indicate
which key has been pressed.  For the Return key, the key code is the
constant |games.K_RETURN| that the library provides for us.

Now that we know how to find out to look into a box, we then have to
decide what to do about it.  The simple answer is to ask the box to do
it for us --- after all, it knows what it is!  What we will need to
know is whether or not we just stepped on a mine, which is something
that the box can return to us.  If we did trigger a mine, we should
end the game, and we already know how to do that!

All this comes down to adding the following function to the
|Minesweeper| class:

\begin{program}
\C{Indent to match the others}
  def handle_keypress(self, key):
    if self.game_over:
      return
        
    if key == K_RETURN:
      if self.cursor.reveal_box():
        self.end_game()
\end{program}

The |cursor| that is used above is something that the library keeps up
for us.  It gives us the box that the cursor is currently over on the
screen, which we can then treat like any other box.

Having done this, the next thing to do is to write the |reveal_box|
function that we have just asserted that the boxes will have.  Clearly
mines and empty boxes will behave differently, so we should write
different functions for the |Mine| and |EmptyBox| classes.

For a mine, if it's stepped on we need to return a ``true'' value
(i.e. anything that isn't zero).  We would also like to explode the
mine and make it visible, but just consider for a moment what is going
to happen.  When we return to the |handle_keypress| function that
we've just written, it is going to end the game and show all the
mines.  If we take any special action to show our own mine, it will be
done all over again, which is a bit of a waste.  In fact, we can just
let the |end_game| do all the work for us!

There is one last little wrinkle here; what do you do if there is
already a flag on the mine?  We could just ignore the flag and explode
anyway, but putting a flag down normally means that the player thinks
that there is a mine here.  It's quite likely that the player pressed
Return by mistake, and meant to do something else entirely.  It's no
trouble for us to be helpful and pretend that nothing happened if
there is a flag there.

In other words, all we have to do is to return a true value if there
is no flag (i.e. if |self.flagged| is false), and to return a false
value otherwise.

\begin{trythis}
You should be now to be able to write the |reveal_box| for |Mine|
yourself.  Here's a hint that may make your function even shorter: if
you use |not x| you get a value that is true when |x| is false, and
false when |x| is true.
\end{trythis}

It's a little trickier to write |reveal_box| for an empty box.  There
are a few decisions we can make quickly.

\begin{itemize}
\item If the box has already been revealed, there is nothing to do.
\item If there is a flag on the box, we should do nothing just like we
  do for a mine.  Otherwise we are giving things away, and giving the
  player a particularly easy way to cheat.
\item We should set |self.shown| to a true value, so that we will know
  if we try to reveal this box again.
\item It would nice to change the background colour of our box if we
  do show it, just to make it more visible.
\item When we are done, we need to return false (0) to show that we
  haven't revealed a mine.
\end{itemize}

A more difficult challenge is presented by the description of the
game.  According to that, we need to show how many of the surrounding
squares contain a mine.  Fortunately the library helps us here by
providing a list of the surrounding boxes in |neighbours|.  After
that, all we need is some way of asking each neighbour whether or not
it is a mine.  We can write a simple function for that, which will
allow us to count up the mines.

Once we have the number, we still need to display it.  The library
|Text| object provides a way to display a string, but it won't display
numbers.  Fortunately, the standard Python function |str| converts a
number into a string for us.

\begin{program}
\C{Type this in the EmptyBox class}
\C{Remember to align the defs!}
  def reveal_box(self):
    if self.shown or self.flagged:
      return 0

    count = 0
    for n in self.neighbours:
      if n.is_mine():
        count = count + 1

    self.set_colour(colour.white)
    self.image = games.Text(self.screen,
                            self.xpos() + BOX_SIZE/2,
                            self.ypos() + BOX_SIZE/2,
                            str(count),
                            NUMBER_SIZE,
                            colour.blue)
    self.shown = 1
    self.board.revealed_box()
    if count == 0:
      for n in self.neighbours:
        n.reveal_box()
    return 0
\end{program}

\begin{trythis}
You still have to write the |is_mine| function for mines and empty
boxes.  That's up to you; just remember that |1| means true and |0|
means false!
\end{trythis}

If you've been paying attention, you'll notice that we've sneaked two
extra things into the code above without telling you.  First, there is
a call to |self.board.revealed_box|, which will be a function in the
|Minesweeper| class that keeps track of how many boxes have been
looked into.  We'll need that later to work out when the game is
finished.

The definition of |revealed_box| is very simple:

\begin{program}
\C{Add this to the Minesweeper class}
\C{Don't forget to indent it!}
  def revealed_box(self):
    self.unknown_count = self.unknown_count - 1
\end{program}

The second thing is just something to make life easier for players.
If there are no mines in the surrounding boxes (i.e. if |count| is
zero), then it is safe to look in to all of them.  We do that on the
player's behalf so that he doesn't have to do it manually for each of
them.

There is only one thing we need to do to be able to run this now; we
have to replace the |minefield.end_game()| instruction with something
that will actually run our program:

\begin{program}
minefield.mainloop()
\end{program}

There, that was easy!

\subsection{Mouse Hunt}

While racing around the window using the keyboard is all very well, it
might be easier to use the mouse to play Minesweeper with.
Fortunately the libraries make this fairly easy for us.

Just like with keypresses, when the library sees a mouse click that it
doesn't understand (which is to say most of them), it calls a function
that we can override.  In this case we'll use the |mouse_up|
function, which is called when the player releases the mouse button.
The libraries give us two parameters to this function; first the
coordinates of where on the screen the mouse was when we let go, and
second the button that was let go.  The buttons are numbered from 0 to
2, with 0 being the left button, 1 the middle button, and 2 the right
button.

If you think about how you normally use the mouse on the computer, you
normally use the left button to select things.  Whether it's a file to
open, some text to cut or copy, or part of a picture to apply an
effect to, the left button is the one we use to select it.  In our
game, the nearest idea we have to that is looking into a box to see
what's there, the same thing that the Return key does.  So what we
need to do is to work out which box the mouse is pointing at (assuming
it is pointing to any box at all), move the cursor there and do the
same for button 0 as we do for the return key.

\begin{program}
\C{Add this to the Minesweeper class}
\C{Get the indentation right!}
  def mouse_up(self, (x, y), button):
    if self.game_over:
      return

    i,j = self.coords_to_cell(x, y)
    if not self.on_board(i, j):
      return
    self.move_cursor(i, j)
    if button == 0:
      if self.cursor.reveal_box():
        self.end_game()
\end{program}


\section{Flying the Flag}

So far we can look into boxes for mines, but we can't plant flags on
them so we still can't win the game.  So what steps do we need to take
to do that?

First, we need to extend our |handle_keypress| function to cope with
an extra keypress to mean ``plant a flag.''  Arbitrarily, we pick the
space bar for this one, so we add the following immediately after the
the |if| statement dealing with the Return key:

\begin{program}
\C{Line up the ``elif'' with the ``if''}
    elif key == games.K_SPACE:
      self.cursor.toggle_flag()
\end{program}

\begin{trythis}
You will want to do something very similar for the mouse too.  Try
adding making the right button (button 2) do the same as pressing the
space bar.
\end{trythis}

Now, if the box has already been shown to the player, then there is
no point in trying to put a flag on it because the player knows it
isn't a mine.  We can assume that he hit the button by mistake.

To put a flag on the box, we will need to draw the shape of the flag
on the screen in the right place.  We can get the coordinates of the
box from the functions defined in the library, so that's easy.  The
hard bit is the flag shape, which is a set of coordinate pairs.  We
have defined something to be going on with in the constant |FLAG| a
little earlier.

Thinking ahead a little, we are going to have to keep track of the
number of flags that we have.  Clearly no one box has any idea what
the other boxes are doing, so we will need to palm this off on the
|Minesweeper| class that contains all the boxes.  For now, we'll just
define what we're going to do later; in particular we will say that
that we will write a function to add or subtract from a total number
of flags displayed that we will have to keep somehow.

If there is already a flag in the box, we should remove it if we are
asked to.  The player may have put it down by mistake, after all.  To
do this we need to remove the flag object from the screen by calling
its |destroy| method (a ``method'' is another name for a function that
belongs to a class rather than existing on its own).

Notice that in all of this, it doesn't matter whether the box we are
flagging is a mine or not.  This suggests that our |toggle_flag|
function should go into our |GridBox| class, since it applies to both
|Mine| and |EmptyBox| types of boxes equally.

The following code will do all this for us.  Be careful when you type
it in that the |def| that it starts with lines up with the |def| for
the |__init__| function in |GridBox|.

\begin{program}
\C{Align with previous def!}
  def toggle_flag(self):
    if self.shown:
      return
    if self.flagged:
      # We have already flagged here.  Remove it
      self.flag.destroy()
      self.flag = None
      self.screen.update_flag_count(-1)
    else:
      # Not yet flagged, so make one
      self.flag = games.Polygon(self.screen, self.xpos(), self.ypos(),
                                FLAG, colour.green, filled=1)
      self.screen.update_flag_count(1)
    self.flagged = not self.flagged
\end{program}

Updating the count of flags is in fact easy.  Just type the following
method into |Minesweeper|:

\begin{program}
\C{Again, it must line up}
  def update_flag_count(self, adjustment):
    self.flag_count = self.flag_count + adjustment
\end{program}

This will add one if we give it a parameter of |1|, and subtract one
if we give it a |-1|, just like we want.  Try it and see!

\begin{trythis}
The flag isn't a very pretty shape at present.  Why don't you
experiment with the coordinates we put in |FLAG| right at the
beginning and see if you can make something that looks more
appropriate?
\end{trythis}


\section{How To Win}

We still have the problem that we don't know when the player has won.
Since players like winning, we had better fix that.

Thinking about it, the player will have won if all the mines are
correctly flagged and the player knows it.  In other words, all the
non-mine boxes must have been looked into, and all the other boxes
must be flagged.

Looking back, we have been keeping track of two things in the
|Minesweeper| class.  First, we have kept track of the number of flags
that are showing in the variable |flag_count|.  Clearly, this needs to
be the same as the number of mines we were looking for.  The second
thing that we have been watching is the number of boxes whose contents
are unknown, |unknown_count|.  Clearly again, the only unknown boxes
should be those with mines in them.  When both of these conditions are
true, all the unknown boxes are flagged and contain mines, and there
are no other mines (or we would have already lost).

This leads us to a nice simple function in |Minesweeper| for checking
if we're done:

\begin{program}
\C{Remember to align the defs!}
  def check_for_win(self):
    if self.flag_count == self.mines_to_find == self.unknown_count:
      self.game_over = 1
\end{program}

Then we just need to call this function (as |self.check_for_win()|)
every time that the player does something to change the board,
i.e. after pressing a key or clicking on the mouse.

\begin{trythis}
Well, what are you waiting for?
\end{trythis}


\section{Telling The Player}

Now we have a working game of Minesweeper.  There's just one problem;
the player doesn't know that he's won or lost, or how many mines he
hasn't found yet.  What we need is some way of getting messages out.

The games library provides us with a means of drawing text on the
screen, as we saw when we were writing the number of mines surrounding
a box earlier on.  We can use |Text| objects to show our useful
information as well, though we need to put them somewhere slightly off
the board so that they don't get in the way of anything important.

We really need two places for messages, one at the top of the screen
to tell the player when he has won or lost, and one at the bottom to
keep track of how many mines have been put down.  Both of them need to
be added to the |Minesweeper| class |__init__| method.  Here's the
first one:

\begin{program}
\C{Remember to align this}
    x,y = self.cell_to_coords(n_cols/2, -1)
    self.status = games.Text(self, x, y + BOX_SIZE/2,
                             "M I N E S W E E P E R",
                             TEXT_SIZE,
                             colour.white)
\end{program}

This puts the message off the top of the board area (in row ``-1'', if
such a thing existed), with its middle point halfway across the board.

\begin{trythis}
Can you work out how to set up a message telling the player how many
mines are yet to be found?  Put it off the bottom of the grid (which
has rows numbered from |0| to |n_rows-1|) and store it in a variable
called |self.mine_count|.

You'll need to turn the number of mines to be found (|mines_to_find|)
into a string and append it to a few words of explanation --- do you
remember how to do that?
\end{trythis}

Once you have created the |mine_count| message, you will need to keep
it up to date.  We must change the number every time we add or remove
a flag, but fortunately we already ensure that |update_flag_count| is
called every time the number of flags changes.  So all we need to do
is to add the following line to that function:

\begin{program}
\C{Remember to line it up}
    self.mine_count.set_text('Mines remaining: ' +
                             str(self.mines_to_find - self.flag_count))
\end{program}

\begin{trythis}
We also need to change the status message to congratulate the player
when he wins and commiserate when he loses.  Can you work out how to
do that?
\end{trythis}


\section{Making Things Quicker}

We have a working game now, but there are a couple of things we can do
to make life a little easier for the player.  The main extra trick
that other versions of Minesweeper have is the ability to reveal
what's in surrounding boxes when the player has flagged all the
surrounding mines, or at least thinks he has.

Let's pause a moment to work out what we need to do for this:

\begin{itemize}
\item If the player hasn't yet looked in the box, we should ignore any
  request to do this.
\item We then need to count up the number of flags and the number of
  mines surrounding the box.  If they aren't the same, we shouldn't go
  any further because the player hasn't counted properly.
\item If we have got the right number of flags, we show the contents
  of each neighbour that isn't flagged.  If it's a mine, the game is
  over because the player made a mistake.
\end{itemize}

When you look at this list, everything is done from the point of view
of the box.  It doesn't actually matter whether the box is empty or
not, although obviously if it is a mine then it won't have been looked
into yet so we will stop at the first hurdle.  These things all
suggest that we should do all the work in our |GridBox| class, the one
that handles all boxes.

The first part is easy.  Add this to the bottom of the |GridBox| class
definition:

\begin{program}
\C{Indent to line up with the other defs}
  def reveal_known(self):
    if not self.shown:
      return 0
\end{program}

We're thinking ahead a little by returning 0 (false) here.  If you
look at our third bullet point above, we might have to end the game if
the player was wrong, so we need to return an appropriate value to
tell the rest of the program.  The last time we did this, for
|reveal_box|, we returned 1 for an exploded mine and 0 for a safe
choice.  For consistency's sake, we should do the same here.

Next, we need to go through all our neighbours and count things up.
Remember how to do that?  It's what the |neighbours| list is for:

\begin{program}
\C{Line this up!}
    number_of_flags = 0
    number_of_mines = 0
    for n in self.neighbours:
      if n.is_mine():
        number_of_mines = number_of_mines + 1
      if n.flagged:
        number_of_flags = number_of_flags + 1
\end{program}

Finally, if everything is to our liking, we need to run through the
list of neighbours again and show them all.

\begin{program}
\C{Indent again}
    if number_of_flags == number_of_mines:
      for n in self.neighbours:
        if not (n.flagged or n.shown):
          if n.reveal_box():
            # Oh dear, a flag was in the wrong place
            return 1
\C{Line this up with the first 'if'}
    return 0
\end{program}

All you need to do now is to call our new function when an appropriate
key is pressed, and end the game if it returns a true value.

\begin{trythis}
Go on then!  Pick a key and add another |elif| clause to the function
|handle_keypress|.  You may need to ask a leader the name that the
library uses for an appropriate key, but if you want an ordinary key
then it's easy.  The 'a' key is called |games.K_a|, 'b' is |games.K_b|
and so on.
\end{trythis}

That's it!  You now have a fully functioning game of Minesweeper.


\section{Bells and Whistles}

There are a few things you could do if you wanted to improve your
game.  See if you can figure out how to do any of the following:

\begin{itemize}
\item Currently you can only play one game of Minesweeper.  When it's
  over, you have to start the program again to get another game.  See
  if you can work out how to start a new game once the player presses
  a given key after the end of the game.  Hint: you need to |destroy|
  objects on the screen before they will disappear.
\item Some versions of Minesweeper keep track of how long you take to
  win the game.  See if you can work out how to do this; ask a leader
  about the |Timer| class in the games library.
\item If you're doing both of the above, try keeping a high score
  table.  Don't worry about saving it, just keep the table for the
  time that the program is running.
\end{itemize}
\end{document}
