\newif\ifaspdf
\ifx\pdfoutput\undefined
\aspdffalse % we are not running PDFLaTeX
\else
\pdfoutput=1 % we are running PDFLaTeX
\aspdftrue
\fi
\ifaspdf
	\documentclass[pdftex]{wsheet}
\else
	\documentclass{wsheet}
\fi
\usepackage{rcs}
\usepackage{graphics}
\usepackage[colorlinks]{hyperref}
\RCS $Id: 2-tables.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\usepackage{ulem}
\sheet{2}{Turning the Tables}
\author{Gareth McCaughan}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Gareth McCaughan. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%---------------------------------------------------------------------------
\section{Introduction}
One thing computers are very, very good at is arithmetic.
This sheet will show you how to make the computer test how good
you are at it. (The computer is probably about 2,000,000,000 times
faster than you, but for now we'll only test how accurate you are.)
%---------------------------------------------------------------------------
\section{What you need to know}
To do this sheet, you need to know:
\begin{itemize}
\item
  How to edit and run a Python program: see Sheet~R (\emph{Running Python})
\item
  The basic bits of Python, from Sheet~1
\item
  Some simple arithmetic!
\end{itemize}
%---------------------------------------------------------------------------
\section{Planning it out}
In sheet 1 you typed things in at the ``prompt'' and got immediate results. 
In this sheet you're going to start writing a ``program''; a bunch of stuff 
you want the computer to do all in one go, in the right order, depending on 
other stuff\dots but don't worry, we'll guide you through it and it's not 
that hard.

When you start to think about writing a program, it's helpful to
begin by thinking exactly what will usually happen when the program
runs. It'll probably go something like this:

\begin{interaction}
What's 6 times 7? \T{49}
No, I'm afraid the answer is 42.
What's 3 times 2? \T{6}
That's right -- well done.
\C{And so on, with several more questions\dots}
What's 5 times 9? \T{45}
That's right -- well done.

I asked you 10 questions. You got 7 of them right.
Well done!
\end{interaction}

So, here are some things we need to be able to make the computer do:

\begin{itemize}
\item Choose numbers (at random, preferably)
\item Display a sum
\item Calculate the right answer
\item Get an answer from the person using the program
\item See whether it's right or not
\item Display a ``that's right'' or ``that's wrong'' message
\item Keep count of how many questions were answered right
\item Ask a total of (say) 10 questions, and then stop
\item Display a final message saying how you've done
\end{itemize}

That's quite a lot of things to do, but most of them are quite
easy. We'll do them one at a time.
%---------------------------------------------------------------------------
\section{Random numbers}
Let's start with the first item on that list: choosing the numbers.
Try asking Python this and see what it says:

\begin{interaction}
>>> \T{from livewires import *} \C{This is magic; see the box below.}
>>> \T{random_between(10,15)}
\end{interaction}

What do you think |random_between| does? Try repeating that last
line a few more times until you're sure you understand how it
behaves.

\begin{note}
That first line there is important. You'll need to put it at the start
of every program you write. It lets you use some things we've added to
the basic Python language, to make your life easier.

As well as putting it in programs you write, you should also
type it in when you start a Python session. Otherwise Python
won't understand some of the things you ask it to do.

See Sheet~M (\emph{Modules}) if you want the gruesome details
of what's going on here.
\end{note}

\emph{Challenge}: Make Python print a sum like

\begin{interaction}
What is 7 times 2 ?
\end{interaction}

where the two numbers are generated randomly, between 1 and 10
(inclusive).
%---------------------------------------------------------------------------
\section{Questions and answers}
OK, so now we know how to make Python display the question.
This isn't much use if we can't get an answer from the
\sout{victim} person using the program. Fortunately,
we can. Try this and see what happens:

\begin{interaction}
>>> \T{print read_number()}   \C{Do you need the \emph{print}? Think back to Sheet~1.}
\end{interaction}

What about this?

\begin{interaction}
>>> \T{print 2*read_number()}
\end{interaction}

\emph{Challenge}: Make Python ask for \emph{two} numbers, add them up,
and print the result. Hint: you can do it with a single command, it's just
a bit longer than the ones you've seen so far.

You can change what Python says when asking for a number, if you
like. Try this:

\begin{interaction}
>>> \T{print read_number('What is your favourite number?')}
\end{interaction}

Now: Write a short program that displays a sum using two random
numbers (as above), and then gets the user to type in an answer.
This should just be a matter of combining two things you've already
done, but in a program rather than on the prompt; check Sheet~R if 
you're not sure how to do that. 
Don't worry about making the program check whether the user's
number is right, or anything. Save this program so that you don't
lose it: you'll be needing it again.
%---------------------------------------------------------------------------
\section{Remembering the numbers}
Well, now is the time to worry about making the program check
whether the user's number is right. This is trickier than what
we've done so far, because we need to use those random numbers
\emph{twice}: once when we're saying what sum we want done, and
once when we're working out the right answer.

\emph{Important Principle}: If you're writing a program and you
need to use something twice, \emph{give it a name}.

We discussed names in sheet~1, but here's a brief reminder.

You give things names by using the |=| sign, and after that
you can use the name instead of the thing you named:

\begin{interaction}
>>> \T{thing=1234}
>>> \T{5*thing}
6170
\end{interaction}

\emph{Challenge}: Write a new program that makes a random number and
prints it out 5 times. (This is \emph{not} the same thing as
printing out 5 different random numbers!) You'll need to give
the random number a name.

\emph{Challenge}: Find the program you wrote that prints a sum
and asks for the answer. Change it so that it gives names to the
numbers in the sum. Check that you've done it right by making it
print the sum twice -- run the program and make sure that it
does ask the \emph{same} question twice, not two different
questions.

Make your program do the following:

\begin{enumerate}
\item Choose two random numbers from 1 to 10.
\item Display a multiplication sum involving them.
\item Ask what the answer is.
\item Say what the answer should have been.
\end{enumerate}

You've already done the first three of these; so the only new
thing is the last one. Save the program again!
%---------------------------------------------------------------------------
\section{Right or Wrong?}
The program you've just written could be used for testing
someone's tables, but it seems rather silly to make the human
check all the answers; that's just the kind of boring job computers
do well. So, the next stage is to get the machine to check whether
the answer you gave was right, instead of just saying what it should
have been and leaving you to decide.

To do this, we need a new -- and very important -- idea.
%---------------------------------------------------------------------------
\section{If \dots}

Write a new program that looks like this, and run it.
(The spaces at the start of some of the lines are important!)

\begin{note}
Even though this isn't underlined, you should still type it in.
The underlining thing is only done when there's a mixture of
things you type in and things that the computer says, to help
you see which is which.
\end{note}

\begin{program}
if 1 < 2:
  print 'Something is wrong!'
else:
  print 'Something is right!'
\end{program}

\begin{note}
The reason why the spaces at the start of a line matter
is that Python uses them to decide how much of your program
is controlled by the |if|. Suppose you say

\begin{program}
if 1 > 1000:
  print 'Ouch!'
print 'Boink!'
\end{program}

Then the computer \emph{will} print |Boink!|, because that line
isn't part of the |if|. But if the |print 'Boink!'| line had
spaces at the start like the |print 'Ouch!'| line, then it
\emph{would} be part of the |if|, and therefore wouldn't get
printed.

Most computer languages don't take any notice of space at the
start of a line. This means that they have to solve the problem
in a different way; usually they need something like |end if|
at the end of the |if|, or else they insist that you surround
everything controlled by the |if| in some kind of brackets.
Python's way is easier to read.
\end{note}

Now change the |<| to |>|, and run the program again. Can you guess
what's going on? (In case you haven't already met the symbols in
school: |<| means ``is less than'', and |>| means ``is greater than''.)

\emph{Challenge}: Write a program that asks you for a number
and then prints different things depending on whether the number
is bigger than 100 or not.

For our tables-testing program, of course, what we want to know
isn't whether the number you typed in is \emph{bigger~than} the
correct answer; we want to know whether it's \emph{equal~to} it.

You might expect to do that by writing
|if |\emph{something}| = |\emph{somethingelse}|:| and so on,
but in fact it turns out that to avoid confusion between
(1) using |=| to mean ``is equal to'' and (2) using |=| to
mean ``is the name of'', Python uses different symbols for
those two things. We've already seen that |=| is how you
say ``is the name of'', so ``is equal to'' must be something
different.

In fact, in Python ``is equal to'' is written with \emph{two}
equals signs, like this: |==| .

(If you want to know more about |if| and related things,
see Sheet~C (\emph{Conditionals}).)

\emph{Challenge}: Change the little program you just wrote that
tests whether a number is bigger than 100, so that instead it
tests whether the number is \emph{equal to} 100.

Now you should have no difficulty making your tables-testing program
print a ``yes!'' or ``no!'' message depending on whether you typed in
the right answer to the sum or not. So now you have a program that
tests you on \emph{one} multiplication sum, and then stops. We're
making progress\dots
%---------------------------------------------------------------------------
\section{Over and over and over again}

You may remember the |for| loop, from sheet~1. Whether you do
or not, here's an example of how to use it. Write a little program:

\begin{program}
for x in 1,2,3,4,5,6:
  print 'x is', x
\end{program}

(Do I still need to remind you that \emph{spaces at the starts of lines
matter}?)

\emph{Challenge}: Put that whole program inside another |for| loop,
so that it all happens three times.

If you managed that, find your tables-testing program and do something
similar so that it does everything 10 times.

If you want to know more about how to do things over and over again
in Python, see Sheet~L (\emph{Loops}).
%---------------------------------------------------------------------------
\section{Who's counting?}
A little while ago I was complaining that our program forced the
person using it to check their own answers. We've fixed that now,
but the user still has to count how many answers they've got right.
The computer ought to be able to do that.

Here's a little program to try. It doesn't have much to do with
the tables-tester, but you'll probably find what you have to do
next easier if you try this first.

\begin{program}
odd=1
for x in 1,2,3,4,5,6:
  print 'An odd number:', odd
  odd = odd+2
\end{program}

That last line may look rather strange: how can |odd| be equal to
|odd+2|? Well, remember that I said a little while ago that in Python
|==| means ``is equal to'', and |=| means ``is a name for''. What
the line tells Python to do is: work out |odd+2|, and then call
\emph{that} |odd|.

(In sheet~1 I mentioned that what we're calling ``names'' are often
called ``variables''. You've now discovered why: the things they name
can change, or \emph{vary}. In the program above, |odd| starts off
meaning 1; then it means 3, then 5, and so on.)

OK. Back to the tables-testing.

Add the following lines to your tables-testing program.
You'll have to work out \emph{where} in the program to put each one.
Some of the lines may need some spaces added at the start.

\begin{program}
right = 0
wrong = 0
right = right + 1
wrong = wrong + 1
print 'You got', right, 'questions right and', wrong, 'wrong.'
\end{program}
%---------------------------------------------------------------------------
\section{Improving the program}
If you've done everything correctly so far, you now have a program
that does roughly what I described at the start of this sheet.
There are lots of things that could be made better; if you aren't
fed up of the program yet, you could try some of these:

\begin{enumerate}
\item
  \emph{Improve the layout}. It's a pity that the question and
  the user's answer have to be on different lines, and that there
  are some unnecessary spaces. Have a look
  at Sheet~I which tells you a bit more about ``input and
  output'': that is, making the program ask things or say things.
  You may find that you also need to know some things described
  in Sheet~S (\emph{Strings}).
\item
  \emph{Timing}. It's probably not hard to get 100\% if you don't
  mind taking, say, an hour over each question. But if you get 100\%
  and take less than a second per question, you're doing very well.
  So, make the program time you and report at the end how long you
  took. You'll need to look at Sheet~T (\emph{Time}) for this.
\item
  \emph{Adjustable difficulty}. Some people are very good with
  numbers. They might find being tested on numbers from 1~to~10
  rather boring. Some people are very bad with numbers, and might
  prefer easier questions. Obviously, it's not too hard to change
  the program to use larger or smaller numbers. (Look at the
  program and make sure you can see how to do that.) It might be
  more interesting if the program became a little harder every time
  you get a question right, and a little easier every time you
  get one wrong. (For this to work well, you'd probably need to
  ask more than 10 questions.)
\item
  \emph{Adjustable length}. You might want a quick test, with only
  four questions. Or a long one, with 100 questions, to see how long
  you can stay awake. Make the program begin by asking how many
  questions you want, and then ask that many.

  To do this, you'll need to know about things called ``ranges''.
  They're described in Sheet~L (\emph{Loops}).
\end{enumerate}
%---------------------------------------------------------------------------
\section{What next?}
The next sheet in the series is sheet 3, ``Pretty pictures''.
It's all about graphics: drawing pictures with Python.
%---------------------------------------------------------------------------
\end{document}
