\documentclass{wsheet}
\usepackage{rcs}
\usepackage[colorlinks]{hyperref}
\RCS $Id: A-lists.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\sheet{A}{Lists in Python}
\author{Rhodri James}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Rhodri James. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%------------------------------------------------------------------
\section{Spam, spam, spam}
Lists are collections of things.  You make lists of what you want when
you go shopping, lists of what's been cooked in the canteen, or lists
of where your robots are on the screen.  So how do you make a list in
Python?

\begin{interaction}
>>> \T{menu = ['spam', 'eggs', 'chips', 'beans']}
>>> \T{numbers = [1, 2, 3]}
>>> \T{both = [2, "wombats", "are", 4, "going elsewhere"]}
>>> \T{empty = []}
\end{interaction}

Yes, you just stick square brackets around whatever you want to turn
into a list, even when that something is nothing at all!  That's not
as silly as it sounds, by the way; everything has to start from
somewhere, so why not start a list with nothing on it, and add items
as you think of it.

\begin{note}
	If you are used to other computer languages, you may already
	be thinking that Python's lists sound a lot like arrays.
	Congratulations, they are arrays, mostly.  Keep reading,
	though; there are some little twists in thinking coming!
\end{note}

%-----------------------------------------------------------------
\section{Slice of spam, sir?}
Now we have this list, how do we find out what's on it?  Python
conveniently numbers everything on our list for us, and lets us find
out what each thing is by writing its number in square brackets after
the list's name.  Try typing the following:

\begin{interaction}
>>> \T{menu[1]}		\C{First thing on the list?}
'eggs'			\C{Not 'spam'.  Oops!}
\end{interaction}

In fact Python numbers items on the list from 0, not from 1, so our
spam is actually |menu[0]|.  You can also use negative numbers to read
from the end of the list, rather than the beginning.  The last item on
the list is numbered |-1|, so this works a little more like you would
expect!

\begin{interaction}
>>> \T{menu[-2]}
'chips'
>>> \T{numbers[-1]}
3
\end{interaction}

Something else you can do with lists is to ``slice'' them, making a
new list from part of the old one.  Just tell Python where you want it
to start taking things from and where you want it to stop, with a
colon |:| in between them.

\begin{interaction}
>>> \T{menu[1:3]}
['eggs', 'chips']	\C{Take items 1 & 2, stop at 3.}
>>> \T{menu[:2]}
['spam', 'eggs']	\C{Start at the beginning if you don't say.}
>>> \T{menu[2:]}
['chips', 'beans']	\C{End at the end if you don't say.}
>>> \T{menu[:]}
['spam', 'eggs', 'chips', 'beans']	\C{Silly!}
\end{interaction}

We can do a few other things with lists that you haven't seen yet.

\begin{interaction}
>>> \T{menu + numbers}	\C{``Concatenation'' works just like strings.}
['spam', 'eggs', 'chips', 'beans', 1, 2, 3]
>>> \T{['x'] * 3}	\C{``Repetition'', again just like strings.}
[ 'x', 'x', 'x']
>>> \T{numbers.append(100)}  \C{Add something to the end of the list.}
>>> \T{print numbers}		     \C{It didn't print out for itself!}
[1, 2, 3, 100]
>>> \T{len(numbers)}	\C{How long is my list?}
4
>>> \T{numbers.reverse()}    \C{Turn the list upside-down.}
>>> \T{print numbers}		     \C{Again, it didn't print for itself.}
[100, 3, 2, 1]
>>> \T{numbers.sort()}	\C{Sort the list into order.}
>>> \T{print numbers}
[1, 2, 3, 100]
>>> \T{numbers.index(3)}\C{Where in the list is 3?}
2                       \C{Position 2 (counting from 0)}
>>> \T{numbers[2]}      \C{Check that position 2 really has the right thing}
3                       \C{Yes!}
>>> \T{numbers.index(1234)}\C{What if it's not there?}
Traceback (innermost last):
  File "<stdin>", line 1, in ?
ValueError: list.index(x): x not in list
\C{Eeeek. So don't do that, then.}
>>> del numbers[2]      \C{Removing an item from the list}
>>> numbers
[1, 2, 100]             \C{It's gone.}
>>> \T{range(3)}	\C{Making useful lists.}
[0, 1, 2]
>>> \T{range(1,4)}	\C{A bit like slicing in reverse.}
[1, 2, 3]
\end{interaction}

(Some of those probably look rather weird -- for instance,
|numbers.sort()|. Don't worry about it!)

The |range| function is very useful for |for| loops --
see Sheet~L (\emph{Loops}).

There are lots more things that you can do with lists; you'll discover
some of them on the holiday!

%----------------------------------------------------------------
\section{Lists of lists of lists of \dots}
Sometimes one list isn't enough.  Suppose you wanted to keep a list of
what you had eaten for breakfast; it's easy enough to write

\begin{interaction}
>>> \T{breakfast = ['coffee', 'corn flakes', 'toast', 'marmalade']}
\end{interaction}

and carry on.  But suppose you wanted to list what you had eaten for
breakfast every day, and you don't always eat the same things.  What
do you do then.

Fortunately, Python is very helpful about this.  Remember that we said
earlier that lists were just collections of things.  Well, lists are
things too, so making lists of lists is just like making lists of
anything else!

\begin{interaction}
>>> \T{breakfast = [}
...   \T{'Monday', ['coffee', 'rice crispies'],} \C{Remember the spaces!}
...   \T{'Tuesday', ['orange juice', 'toast', 'marmalade'],}
...   \T{'Wednesday', ['coffee'] ]}
\end{interaction}

(Look, I was in a hurry on Wednesday!)

\begin{note}
	Some people call lists of lists ``2-dimensional arrays'' or
	``tables'', because you can write them out in rows and columns
	(two dimensions!) like a table.  Unlike real tables (and many
	other computer languages), in Python you don't have to have
	every row the same length.
\end{note}

To get at the list items, just add ``indices'' (item numbers) in
square brackets to the end of the list name.  For a list in a list you
need two sets of square brackets.  For a list in a list in a list you
need three, and so on.

\begin{interaction}
>>> \T{breakfast[3]}
['orange juice', 'toast', 'marmalade']
>>> \T{breakfast[3][0]}
'orange juice'
\end{interaction}

%----------------------------------------------------------------
\section{A subtle trap}
So far, we've talked about lists as collections of things.  Actually,
it's more accurate to say that Python's lists are really collections
of ``references'' or ``pointers'' to things.  For most purposes, this
makes no difference at all, and you can happily ignore us being
pedants here.  However, when you make lists from variables or other
lists, then it can become important.

For example, when you replicate a list, you get copies of the same
\emph{identical} list.  They are all actually the same list, so when
you change one, you actually change them all.  So the following
doesn't work quite the way you might have expected:

\begin{interaction}
>>> \T{wombat = [[1, 2, 3]] * 3}
>>> \T{print wombat}
[[1, 2, 3], [1, 2, 3], [1, 2, 3]]
>>> \T{wombat[0][2] = 4}
>>> \T{print wombat}
[[1, 2, 4], [1, 2, 4], [1, 2, 4]]
\end{interaction}

There are ways around this; the easiest is do something that makes a
new list from the oldest, such writing the list out separately each
time or (if you are copying from a variable) taking a slice of the
whole list.  Maybe typing |numbers[:]| isn't so silly after all!

(If you didn't understand any of that last bit, don't worry.)

%----------------------------------------------------------------
\end{document}
