\documentclass{wsheet}
\usepackage{rcs}
\usepackage[colorlinks]{hyperref}
\RCS $Id: D-dicts.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\sheet{D}{Dictionaries and tuples}
\author{Gareth McCaughan}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Gareth McCaughan. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%-----------------------------------------------------------------------------
\section{Introduction}
%-----------------------------------------------------------------------------
The worksheets that make up most of this course don't cover everything.
Two important things they don't cover are ``dictionaries'' and ``tuples''.
This sheet tells you a little about those.

\section{Dictionaries}

The easiest way to explain what a dictionary is is to give an
example. So here is one.

\begin{interaction}
>>> \T{dict = \{\}}       \C{An empty dictionary.}
>>> \T{dict[1] = 'boo!'}  \C{Add a piece of information to it.}
>>> \T{dict['ouch'] = 99} \C{And another.}
>>> \T{dict}              \C{Lets's see what's in there.}
{'ouch': 99, 1: 'boo!'}   \C{Two ``associations''.}
>>> \T{dict['ouch']}      \C{What's associated with |'ouch'| ?}
99                        \C{This is.}
\end{interaction}

So, a dictionary is a little like a list, except that while an
array has entries labelled by numbers (|a[0]|, |a[1]|, etc), the
entries in a dictionary can be labelled by things other than
numbers. (Most usefully, they can be labelled by numbers and
by strings.)

It's sometimes helpful to think of a dictionary as being made up
of ``associations''. An association is a piece of information
like ``|'ouch'| is associated with |99|''. So the dictionary we
built above contains two associations.

You might use a dictionary to hold an address list, with each
association pairing a person's name up with their address.
Or it might hold information about all the players in a
multi-player game, or about all the computers connected
to a network.

The two parts of an association are sometimes called the \emph{key}
and the \emph{value}: you find the value by looking up the key. So
in the dictionary we made above, the keys are |1| and |'ouch'|,
and the corresponding values are |'boo!'| and |99|.

As we've seen, you use dictionaries in the same sort of way
as you use lists: if |d| is a dictionary then |d[k]| is the
value associated with the key |k|.

To change an association in a dictionary, just say something
like |dict['ouch'] = 1000|. To remove an association entirely,
say |del dict['ouch']|. (|del| is short for ``delete''.)

To test whether a dictionary has any association with key |k|,
do something like this:

\begin{interaction}
>>> \T{dict = \{'a': 123, 'b': 987\}} \C{Set up a dictionary}
>>> \T{dict.has_key('a')}             \C{Any association for |'a'| ?}
1                                     \C{Yes.}
>>> \T{dict.has_key('z')}             \C{Any association for |'z'| ?}
0                                     \C{Nope.}
\end{interaction}

\subsection{Keys}

Not all objects are allowed as keys for dictionaries. We've
already seen that numbers and strings are OK. Unfortunately,
lists aren't. However, there's a data type very similar to
the list that you \emph{can} use: the ``tuple''. Tuples are
the other subject of this sheet\dots

\section{Tuples}

A tuple is like a list. There are only two differences you need
to care about:
\begin{enumerate}
\item Tuples are ``immutable''. In other words, once you've got a tuple
      you aren't allowed to change its contents. If |x| is a list then
      you can say |x[1]=99|; if it's a tuple, you can't.
\item Tuples are written a little differently: instead of
      |[1,2,3,4]| you say |(1,2,3,4)|.
\end{enumerate}

You can use a tuple almost anywhere where you'd use a list.
For instance, you can say |for x in (1,2,3,4):| instead of
|for x in [1,2,3,4]| (see Sheet~L (\emph{Loops}) if you don't
know what that's about). 

If for some reason you want a tuple containing only one object,
you can't write it as |(x)| (can you think why?). Instead, you
say |(x,)|. Weird, I'm afraid \dots
\end{document}
