\documentclass{wsheet}
\usepackage{rcs}
\usepackage[colorlinks]{hyperref}
\RCS $Id: E-errors.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\sheet{E}{Errors (also called Exceptions)}
\author{Gareth McCaughan}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Gareth McCaughan. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%-----------------------------------------------------------------------------
\section{Introduction}

When something goes wrong, Python responds by giving you a rude
message that might look something like this:
\begin{program}
Traceback (innermost last):
  File "<stdin>", line 1, in ?
  File "<stdin>", line 1, in f
TypeError: illegal argument type for built-in operation
\end{program}

You can usually ignore everything except the last line, although
the earlier lines may -- if you look at them closely -- give some
hints about where the trouble happened.

This sheet is a brief guide to some of the commoner things you might
see on that last line, what they mean and what you might have done
to provoke them.

\section{Error messages}

\subsection{Attribute Errors, Key Errors, Index Errors}

These have messages starting ``|AttributeError:|'', ``|KeyError:|''
or ``|IndexError:|''. They all mean something rather similar: you
were trying to get at a part of an object (one element of a list,
for instance), but you asked for a non-existent part of the object.
So, if |x| is the list |[1,2,3]| and you ask for |x[100]| you'll
get an |IndexError|.

\subsection{Name Errors}

These all have messages starting ``|NameError:|''. They mean
``I've never heard of this thing''. The ``thing'' Python's
never heard of is what comes after ``|NameError:|''. This
might mean that you mistyped something: |primt x| instead of
|print x|, say. It might also mean that you used a variable
(i.e., a name) before it was defined. Or that you used a
\emph{function} before it was |def|ed.

One way this can sometimes happen is if you forget the quotation marks
around a string.

\subsection{Syntax Errors}

These all have messages starting ``|SyntaxError:|''. What they
have in common is that you said something Python couldn't even
begin to make sense of. If someone says ``I apple eat like the to''
then that's a syntax error!

|SyntaxError: invalid syntax|

This can mean lots of things. Here are some examples of things
that provoke it.

\begin{program}
+ + +          \C{Add \emph{what} to \emph{what}?}
1z6            \C{Is it a number? Is it a variable? Is it a mistake?}
for            \C{After |for| there ought to be some more stuff}
while if:      \C{|if| isn't a thing that can be true or false!}
def 1(x):      \C{|1| is a number. You can't use it as a name for a function}
f([)           \C{The |[| starts a list, but you never finished it}
if 1==2        \C{You missed out the ``:'' at the end of the line}
if x=y:        \C{That should say |==|, not |=|}
\end{program}

One way in which you can get slightly surprising syntax errors is
if you mess up the indentation (spaces at starts of lines), so if
you can't find anything else wrong it's worth checking that the
indentation is right.

|SyntaxError: invalid token|

This usually means that you messed up when typing a string.
Maybe you missed off the final quotation mark, or tried to
put apostrophes into a string surrounded by single quotes,
or something like that.

\subsection{Type Errors}

These all have messages starting ``|TypeError:|''. What they
have in common is that Python was expecting one kind of object
(a number, maybe) and you gave it another (a string, or a list,
perhaps).

A lot of these things can happen in non-obvious ways. For instance,
if you hand something that isn't a number to one of the graphics
functions -- |move()|, |draw()| etc -- then you are likely to
get a |TypeError|.

|TypeError: illegal argument type for built-in operation|

\emph{Means}: You asked Python to do a ``built-in operation''
(i.e., something like |+|) on the wrong sort of object. For
instance, you might have asked it to add a number to a string.

|TypeError: len() of unsized object|

You asked for the length of something that doesn't have a length.
|len()| makes sense for strings, lists and tuples (you don't need
to know what a ``tuple'' is), and not much else.

|TypeError: not enough arguments; expected 1, got 0|

You tried to use a function, but the function was defined with more
arguments than you gave it. You probably forgot one!

|TypeError: number coercion failed|

You tried to do something using numbers, but one of the things involved
wasn't a number. You can get this by doing |4+'ouch'|.

|TypeError: too many arguments; expected 1, got 3|

You tried to use a function, but the function was defined with fewer
arguments than you gave it.

|TypeError: unsubscriptable object|

You tried to do something like |a[3]| where |a| was the wrong kind
of thing. This can easily happen if you use a function that expects
a string and give it something else instead.

\end{document}
