\newif\ifpdf
\ifx\pdfoutput\undefined
\pdffalse % we are not running PDFLaTeX
\else
\pdfoutput=1 % we are running PDFLaTeX
\pdftrue
\fi
\ifpdf
	\documentclass[pdftex]{wsheet}
\else
	\documentclass{wsheet}
\fi
\usepackage{rcs}
\usepackage[colorlinks]{hyperref}
\RCS $Id: F-functions.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\sheet{F}{Functions}
\author{Gareth McCaughan}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Gareth McCaughan. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%-----------------------------------------------------------------------------
\section{Introduction}

A \emph{function} is just a piece of Python program with a name.
Functions are very important; this sheet tells you all about
functions in Python.

\section{What's a function?}

A function is like a recipe. It's a set of instructions that
you can refer to by name. Using functions saves thinking,
in the same sort of way as using recipes does; if you're
planning a meal it's much easier if you can just say ``We'll
have spaghetti bolognaise and apple crumble'' rather than
having to plan every detail of the meal ``from first principles''.

Functions are useful when you want to do the same thing several
times: you just define a function once (like writing down a
recipe) and then use it several times (like making several
cakes with the same recipe).

They're even more useful when you want to do \emph{almost} the
same thing several times. You can write a function so that some
bits of what it does aren't completely specified, and when
the function is used the details get filled in. This is like
having a recipe that will make several different kinds of pie,
I suppose.

Functions are also useful even if you're only going to use
each function once. They don't make your program any shorter
then, but they do make it easier to understand. (You'll probably
just have to take my word for this, until you've written lots
of programs and discovered it for yourself!)

\section{Defining and using functions}

To define a simple function (that does the same thing every
time you use it), do something like this:

\begin{program}
def some_name_or_other():
  do something
  do something else
  blah blah blah
\end{program}

And to use it, you just say

\begin{program}
some_name_or_other()
\end{program}

That will do basically the same as copying everything inside
the function definition would have done.

\section{Local variables}

Suppose you say

\begin{program}
def say_boo_twice():
  boo = 'Boo!!!'
  print boo, boo

boo = 'boo hoo'
say_boo_twice()
\end{program}

Then after doing that, |boo| will still be a name for the string
|'boo hoo'| -- \emph{not} for |'Boo!!!'|. (So this is something
that's different from what would happen if you just copied the
definition of the function instead of ``calling'' the function
by its name.)

There's a good reason for this. It means that functions can't have
terribly unexpected effects. Someone reading the second half of that
little program wouldn't have any reason to suspect that |say_boo_twice()|
would change the value of |boo|. So, making changes to variables ``local''
to the function makes programs easier to understand: you don't need to
know all about what's inside |say_boo_twice()| in order to understand
a piece of program that uses it.

The variable |boo| inside |say_boo_twice| is called a ``local variable''.

\section{Global variables}

Actually, I lied to you when I said that calling a function can't
have unexpected effects on variables' values. It can, but you have
to ask for this to happen. If we change the program above by adding
one extra line

\begin{program}
def say_boo_twice():
  global boo         \C{This is the extra line}
  boo = 'Boo!!!'
  print boo, boo

boo = 'boo hoo'
say_boo_twice()
\end{program}

then the value of |boo| \emph{will} change. Saying |global boo| means:
``The variable |boo| is not a local variable.'' (``Global'' is the
opposite of ``local''.)

You shouldn't need to do this very often, especially because of a
useful rule: If a function uses a variable \emph{but never changes its
value}, then the variable is assumed to be global. So your functions
can \emph{use} variables from the rest of your program without the
trouble of writing |global|; it's only if you want to change those
variables that you need to say they're |global|. And, usually, you
shouldn't need to do that.

\section{Functions with arguments}

The most interesting functions have ``arguments''. An argument is
a piece of information you give when using a function that makes
a difference to what the function does. If you have a recipe that
tells you how to make any number of pancakes (``allow a pint of milk
for every 12 pancakes'', etc) then the number of pancakes is an
\emph{argument} to the recipe (though cookery books don't usually
put it that way!).

Here's how you define a function with arguments.

\begin{program}
def print_two_things(x,y):
  print 'x is', x
  print 'y is', y
\end{program}

You can probably guess that if you say |print_two_things(1,'boing!')|
then the computer will say

\begin{program}
x is 1
y is boing!
\end{program}

(The arguments to a function are really just a special kind of local
variable: inside |print_two_things|, |x| and |y| are local variables
whose values are given when you say |print_two_things(1,'boing!')|
or whatever.)

\section{Returning values}

Some functions don't just \emph{do} things; they also produce
results. For instance, you can write a function that takes two
arguments and has, as its result, the sum of those two arguments.
(There's no particular reason why you'd want to do that; it
just makes a simple example.)

Here's how you do that.

\begin{program}
def add(a,b):
  return a+b
\end{program}

(The magic word |return| means ``the following is the result
of the function''.) So, you could use this function like so:

\begin{program}
print '3 plus 4 is', add(3,4)
\end{program}

which would, unsurprisingly, print |3 plus 4 is 7|.

\section{Optional arguments}

Sometimes the arguments to a function are \emph{usually} the same,
but occasionally you want to do something different. You can save
some typing by using ``optional arguments''. Here's how to do that.

\begin{program}
def do_something(x, y=999):
  print x,y
\end{program}

You can now use this in three ways.

\begin{enumerate}
\item As an ordinary function with two arguments. |do_something(5,6)|
      does just the same as if the definition had just begun with
      |def do_something(x,y):| .
\item As a function with \emph{one} argument. If you don't say what
      value |y| should get, it will be set to 999 for the duration of
      the function. So, |do_something(1)| is the same as
      |do_something(1,999)| .
\item As a function with one argument, and a ``keyword argument'' which
      you can call by name. In other words, you can say
      |do_something(8, y=123)|. Obviously this isn't very
      useful, but if |do_something| had had \emph{several}
      optional arguments this lets you give special values
      to some of them but not others.
\end{enumerate}

\section{Weird stuff}

There are some slightly strange things you can do with functions.
For instance, as far as Python is concerned, a function is just
another object, no more unusual than a number or a string. So you
can use one function as an argument to another. For instance,
the built-in function (i.e., a function Python provides for you
without you having to define it) called |map| works like this:

\begin{interaction}
>>> \T{def thrice(x):}
... \T{  return 3*x}
...                     \C{Just hit \key{Enter} here.}
>>> \T{map(thrice, [1,2,3,100])}
[3, 6, 9, 300]
\end{interaction}

You might find a use for this some day.

\section{Why bother?}

It's possible to write quite complicated programs without
bothering with functions at all. What does using functions
gain you?

\begin{itemize}
\item \emph{Your programs will be clearer}. It's easier to understand
      a program when it's divided up into manageable portions,
      and functions are a useful way of doing that.
\item \emph{You won't have to think so hard}. If your program is
      divided up sensibly into functions, you can understand what
      each bit of it does without having to remember the details
      of every other bit. When you see |update_high_scores()| you'll
      know what that does, without the effort of reading through the
      code that actually updates the high scores, so it will be
      quicker to understand the bit of the program that uses
      |update_high_scores()|.
\item \emph{You'll save typing}. If you have to do the same thing
      (or even roughly the same thing) several times, put it in a
      function. Then you only have to explain how to do it once.
\item \emph{Your programs will be more adaptable}. If you find
      that you need to change something about your program, it's
      much easier to do that if it's neatly packaged up into
      functions. Otherwise you'll have nightmares wondering
      whether you really caught \emph{every} place in the program
      that deals with the player's score and adjusted it for the
      new scoring rules.
\item \emph{You'll be able to re-use more of what you write}.
      With a bit of luck, you'll find that some functions you
      write are useful for several things, so that you can use
      the same function (or a slight variant) in several programs.
      That saves you the effort of having to write it several
      times!
\end{itemize}

You probably won't see the point of lots of the things on this list
until you've actually written (or read!) some complicated programs.
But once you've done that, you'll discover that functions are
essential for keeping large programs under control.
%-----------------------------------------------------------------------------
\end{document}
