\documentclass{wsheet}
\usepackage{rcs}
\usepackage[colorlinks]{hyperref}
\RCS $Id: G-graphics.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\sheet{G}{Graphics}
\author{Gareth McCaughan}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Gareth McCaughan. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%-----------------------------------------------------------------------------
\section{Introduction}

This sheet describes briefly all the graphics things (i.e., drawing
pictures) you can easily do with Python. (You can actually do lots
of other complicated things, but they're, errrm, complicated. We've
made a bunch of useful things easy, and these are what this sheet
describes.)

For any of this stuff to work, you need to have done

\begin{program}
from livewires import *
\end{program}

at some point. If you're writing a program, that should be the first
line. If you're just typing commands one by one into the ``Python Shell''
window, you should type the |from livewires ...| thing once; after that,
you don't need to do it again unless you quit Python.

There's an introduction to graphics in Sheet~3, so this sheet is more
a reference than an introduction. Some of it is \emph{very} terse;
ask a leader if it doesn't make any sense. (If you \emph{are} a leader
and it still doesn't make any sense, ask Gareth.)

\section{Coordinates}

(0,0) is at the bottom left of the window. The window is 640 pixels by 480,
by default. (You can make it a different size if you want to.)
Coordinates are given in units of one pixel.

Almost all functions that take coordinates as arguments can take them
either as two arguments, or as a tuple (x,y). So these two lines
are equivalent:

\begin{program}
circle(300,200,10)
circle((300,200),10)
\end{program}

\section{Starting and finishing}

Before you do anything else, you should say |begin_graphics()|.
This will make a new window; the graphics commands will make things
happen in that window.

After you're finished, you should say |end_graphics()|. The computer
will then wait for the window to be closed.

You can change various things about the window:

\begin{program}
begin_graphics()                      \C{640 by 480, white background}
begin_graphics(width=800,height=600)  \C{A larger window}
begin_graphics(background=Colour.black)\C{Black background}
begin_graphics(title='Walrus')        \C{Different window title}
\end{program}

Once the window is there, though, you can't change it.

\section{Drawing things}

There's always a ``current point''. It starts out at (0,0). Many
drawing commands start drawing at the current point; most also
move the current point.

\begin{tabular}{ll}
|clear_screen()| & Empty the graphics window. \\
|move(x,y)|      & Change the current point. Don't draw anything. \\
|draw(x,y)|      & Draw a line from the current point to (|x|,|y|).
                   Change the current point to (|x|,|y|). \\
|plot(x,y)|      & Plot a single pixel at (|x|,|y|). Update current point. \\
|line(x,y,x,y)|  & Draw a line between two endpoints. \\
|box(x,y,x,y)|   & Draw a box given two opposite corners. \\
|circle(x,y,r)|  & Draw a circle given centre and radius. \\
|text(t)|        & Write some text.
\end{tabular}

Most of these can take an optional ``keyword argument'' specifying
the colour to use: |draw(x,y,colour=red)| .

Some (|box| and |circle|) can take a keyword argument saying whether
to fill them in or not (default is not to fill):
|circle(300,200,100,filled=True)| .

|circle| can draw circular arcs too. Keyword parameter |endpoints|;
should be a list or tuple of length 2, each element being either a
point on the screen or an angle. If the first endpoint given is a
point, you don't need to give the radius of the circle.

There's also |polygon()|. Argument is a list of coordinates.
Keyword arguments: |colour|, |closed|, |filled|.

|text| takes some keyword arguments too. |size|, |serifs|. You
can change the default values of these with |set_textsize(sz)|
and |set_textserifs(x)|.

\section{Colours}

Just as there's a ``current point'', there's a ``current colour''.
By default, it starts off being black. Colours look like this:
|Colour(r,g,b)| where |r|,|g|,|b| are numbers between 0 and 1
indicating how much red, how much green and how much blue there are
in the colour you want. So, for instance, |Colour(0,0,0)| is black;
|Colour(1,1,1)| is white; |Colour(1,1,0)| is bright yellow.

You can assign colours to variables (i.e., give them names):
|yellow = Colour(1,1,0)|, for instance.

To change the current colour, do |set_colour(c)| where |c| is
a colour. This function call returns the ``current colour'' that
was in force before you changed it, so you can do things like

\begin{program}
def draw_red_line():
  old_colour = set_colour(Colour.red) \C{|Colour.red| is pre-defined}
  move(100,100)
  draw(200,200)
  set_colour(old_colour) \C{Leave current colour as it was before we started}
\end{program}

As that program fragment indicates, there are some colours defined
for you: |Colour.red|, |Colour.green|, |Colour.blue|, |Colour.black|
and |Colour.white| . In fact, there are some more: |Colour.|\emph{foo}
where \emph{foo} is one of: |dark_grey|, |grey|, |light_grey|,
|dark_red|, |dark_green|, |dark_blue|, |yellow|, |brown| .

\section{The mouse}

We can detect mouse movement and clicks, but doing so may slow things
down. If you want to do it, say |mouse_begin()| before trying, and
|mouse_end()| when you're done.

\begin{tabular}{ll}
|mouse_position()| & Return current mouse position as a 2-tuple. \\
|mouse_buttons()|  & Return a dictionary mapping |'left'| etc to 0 or 1. \\
|mouse_wait(how)|  & Wait for something to happen according to |how|.
\end{tabular}

Values of |how|: |'down'|, |'up'|, |'change'| wait for suitable
value of button state. |'click'| waits for down then up.
|'move'| waits for the mouse to move. |'any'| waits for a change
in either position or button state.

\section{The keyboard}

You can also tell what keys are pressed; this might be useful for
games. |keys_pressed()| returns a list containing all the keys
currently pressed.

\section{Movable objects}

If you say |allow_moveables()| then from then on, all calls to
|line|, |box|, |circle|, |text| and |polygon| will return
values. These values can be passed to functions that move
the corresponding objects around:

|remove_from_screen(object)| removes the object completely

|move_by(object,dx,dy)| moves the object |dx| pixels to the right
and |dy| pixels upwards. (|dx| and |dy| can be negative if you like.)

|move_to(object,x,y)| moves the object to the position |(x,y)|.

The ``position'' of a circle is the position of its centre.
The position of a box or line is the \emph{first} pair of
coordinates used to create it. The position of a polygon
is the first pair of coordinates in its specification.

If you want to go back to immovable objects, say |forbid_moveables()|.
Incidentally, you can spell ``movable'' either with or without the
`e' in the middle. Both spellings are correct.

\end{document}
