\documentclass{wsheet}
\usepackage{rcs}
\usepackage[colorlinks]{hyperref}
\RCS $Id: M-modules.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\sheet{M}{Modules}
\author{Gareth McCaughan}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Gareth McCaughan. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%-----------------------------------------------------------------------------
\section{Introduction}

Python provides a \emph{huge} number of useful functions and
other such things. (These worksheets describe, perhaps, about 1\%
of what there is. Maybe 2\%.) So many different functions might
be confusing, so they're parcelled up into things called
``modules''.

\section{Importing modules}

You can think of a module as a collection of useful things.
(The actual definition is a bit complicated, and you really don't
need it.) To make the Useful Things in a module available to you,
you have to ``import'' the module. So:

\begin{program}
import stuff
\end{program}

makes the things in the module |stuff| available for your program
to use. Those things will have names that look like |stuff.thingy|
so that there's no danger of two things in different modules
``colliding'' by having the same name. It's a bit like with
telephone numbers: there are probably lots of people whose
phone number is 321987; that's why there are ``dialling codes''
like 01223 (for Cambridge). The module name is like a dialling
code.

\section{Living dangerously}

It can get very tiresome, having to type a module name in
over and over again. So Python gives you a way to get at
the things in a module more directly. If you say

\begin{program}
from stuff import blarp,weeble
\end{program}

then the things that would have been called |stuff.blarp| and
|stuff.weeble| are available under the shorter names |blarp|
and |weeble|. (This doesn't import anything else from the
module, under any name.)

It might be worth doing this if you find yourself using some
bits of a module a lot.

\section{Living even more dangerously}

What if you use \emph{lots} of things from a module a lot?
Then you can say

\begin{program}
from stuff import *
\end{program}

which has the effect of making \emph{everything} (well, almost
everything) in module |stuff| available without needing to
type |stuff.| at the start of any of the names. Doing this
is usually a Bad Idea; there might be two modules containing
different things with the same name, and then all kinds of
bad things could happen.

There's just one case in which we recomend that you do this:
the special LiveWires module we've written to make your life
easier. Any program that uses anything from it should begin
with |from livewires import *| . (The main reason for
that is that we wanted you to be able to use the things in
the module without needing to understand what modules are
and why they result in names like |stuff.thingy| \dots)

\section{Finding out what's in a module}

After doing |import stuff|, you can get a list of all the
names of things in |stuff| by typing |dir(stuff)|. This
probably won't usually be very helpful, but if you're
feeling like exploring \dots

\section{Some useful modules}

As I said at the start, there are a lot of modules out there.
There are more than 170 modules that come as standard with
Python! When you're feeling brave, you might like to find out
about some of them. Here's a very brief list of things it's
worth knowing about. Look them up in the official Python
documentation, or investigate for yourself. (I recommend
looking them up.) To understand some of the documentation,
you may need to know about ``classes'' and ``objects'':
see Sheet~O (\emph{Objects and Classes}).

\begin{tabular}{ll}
|sys|    & Mostly some slightly useful variables. \\
|string| & Things for strings. \\
|re|     & ``Regular expressions'': a complicated but powerful way
           of doing more advanced things with strings. \\
|math|   & Mathematical functions and constants. \\
|random| & Random numbers. \\
|time|   & Time and date. \\
\end{tabular}

%-----------------------------------------------------------------------------
\end{document}
