\documentclass{wsheet}
\usepackage{rcs}
\usepackage[colorlinks]{hyperref}
\RCS $Id: S-strings.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\sheet{S}{Strings}
\author{Gareth McCaughan}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Gareth McCaughan. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

%-----------------------------------------------------------------------------
\section{Introduction}

A ``string'' is a piece of text. Strings are made up of ``characters'';
a character is a single letter, digit, symbol or whatever. Python provides
lots of things for doing with strings; this sheet tells you about some
of them.

\section{Entering strings}

You can put a string into your program by putting it inside
quotation marks. Single quotes and double quotes are both
fine, but you must use the same sort of quotation mark at
the start and at the end!

In these worksheets, I've usually used single quotes. If you
prefer double quotes, that's fine. If you want a string that
contains a single-quote character (which also serves as the
apostrophe), you should surround it with double quotes:
|"I'm bored"| rather than |'I'm bored'| (which won't work
at all: can you see why?)

\section{Strings as sequences}

There are some things that work on lists as well as on strings.
They work because both lists and strings are ``sequences''. Here's
a very brief summary. (You might want to compare with with
Sheet~A (\emph{Lists}).)

\begin{interaction}
>>> \T{thing = 'I am the walrus'}  \C{A string to work with}
>>> \T{gloop = "Another string"}   \C{And another one}
>>> \T{thing+gloop}                \C{String concatenation \dots}
'I am the walrusAnother string'    \C{\quad works as you might guess}
>>> \T{2*thing}                    \C{String replication \dots}
'I am the walrusI am the walrus'   \C{\quad also does what you'd think}
>>> \T{thing[0]}                   \C{We start counting at 0}
'I'                                \C{\quad so we get the first character}
>>> \T{thing[1:5}                  \C{Characters 1 (inclusive) to 5 (exclusive)}
' am '                             \C{\quad so 4 characters in all}
>>> \T{len(thing)}                 \C{How many characters?}
15                                 \C{That many!}
\end{interaction}

\section{The `string' module}

If you say |import string| then after that you can use a whole lot of
functions specially for working on strings. Here are a few useful ones:

\begin{interaction}
>>> \T{import string}              \C{Make this stuff available}
>>> \T{string.capitalize('walrus')}
'Walrus'
>>> \T{string.lower('WalRUs')}
'walrus'
>>> \T{string.upper('WalRUs')}
'WALRUS'
>>> \T{string.split('I am the walrus')}
['I', 'am', 'the', 'walrus']
>>> \T{string.replace('I am a twit', 'a', 'the')}
'I them the twit'
\end{interaction}

(The names, with those dots in, are a bit strange. You might
like to look at Sheet~M (\emph{Modules}).)

There are a \emph{lot} more things you can do with strings if you
want to. If there's something you can't see how to do, ask a leader.

\section{Other things}

It's possible to convert just about any object into a string
that represents it. So, for instance, 1234 would become the
string |'1234'|. The function that does this is called |repr|
(short for ``representation''). So, whatever |x| is, |repr(x)|
is a string describing it.

If |repr(x)| is too clumsy or too much effort to type, you can
instead say |`x`| -- notice that those are a different kind of
quotation mark: |`| isn't the same as |'| !

\end{document}
