\newif\ifpdf
\ifx\pdfoutput\undefined
\pdffalse % we are not running PDFLaTeX
\else
\pdfoutput=1 % we are running PDFLaTeX
\pdftrue
\fi
\ifpdf
	\documentclass[pdftex]{wsheet}
\else
	\documentclass{wsheet}
\fi
\usepackage{rcs}
\usepackage[colorlinks]{hyperref}
\RCS $Id: W-livewires.ltx 239 2010-07-23 21:41:31Z RobPearce $
\RCS $Date: 2010-07-23 22:41:31 +0100 (Fri, 23 Jul 2010) $
\RCS $Revision: 239 $
\sheet{W}{The LiveWires module}
\author{Gareth McCaughan, Paul Wright and Rhodri James}
\date{Revision \RCSRevision, \RCSDate}
\begin{document}

\section{Credits}

% COPYRIGHT NOTICE:
\copyright{} Gareth McCaughan, Paul Wright and Rhodri James. All rights reserved.
% 
% CONDITIONS:
% 
% A "Transparent" form of a document means a machine-readable form,
% represented in a format whose specification is available to the general
% public, whose contents can be viewed and edited directly and
% straightforwardly with generic text editors or (for images composed of
% pixels) generic paint programs or (for drawings) some widely available
% drawing editor, and that is suitable for input to text formatters or for
% automatic translation to a variety of formats suitable for input to text
% formatters. A copy made in an otherwise Transparent file format whose
% markup has been designed to thwart or discourage subsequent modification
% by readers is not Transparent. A form that is not Transparent is
% called "Opaque". 
% 
% Examples of Transparent formats include LaTeX source and plain text.
% Examples of Opaque formats include PDF and Postscript.  Paper copies of
% a document are considered to be Opaque.
% 
% Redistribution and use of this document in Transparent and Opaque
% forms, with or without modification, are permitted provided that the
% following conditions are met: 
% 
% - Redistributions of this document in Transparent form must retain
%   the above copyright notice, this list of conditions and the following
%   disclaimer. 
% 
% - Redistributions of this document in Opaque form must reproduce the
%   above copyright notice, this list of conditions and the following
%   disclaimer in the documentation and/or other materials provided with
%   the distribution, and reproduce the above copyright notice in the
%   Opaque document itself.
% 
% - Neither the name of Scripture Union, nor LiveWires nor the names of
%   its contributors may be used to endorse or promote products derived
%   from this document without specific prior written permission. 
% 
% DISCLAIMER:
% 
% THIS DOCUMENT IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS ``AS
% IS'' AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO,
% THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
% PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDERS,
% CONTRIBUTORS OR SCRIPTURE UNION BE LIABLE FOR ANY DIRECT, INDIRECT,
% INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
% NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
% DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
% THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
% (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF
% THIS DOCUMENT, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

This document is part of the LiveWires Python Course. You may
modify and/or distribute this document as long as you comply with the
LiveWires Documentation Licence: you should have received a copy of the
licence when you received this document.

For the \LaTeX{} source of this sheet, and for more information on
LiveWires and on this course, see the LiveWires web site at
\href{http://www.livewires.org.uk/python/}{|http://www.livewires.org.uk/python/|}

\section{Introduction}

The version of Python you're using comes with a ``package'' called
|livewires|. There are four modules inside the package: |beginners|,
|games|, |boards| and |colour|. |beginners| was produced for the
LiveWires holiday in 1999, though it has been modified since. |games|
and |colour| were produced for the LiveWires holiday in 2000. |games|
was modified for 2001 so that it uses the Pygame library, and has been
tinkered with ever since.  |boards| was written in 2001 while |games|
was being modified.

\emph{Note:} If you've got an old version of the LiveWires package,
containing only the |livewires.py| file, you've got the 1999 version of
the |beginners| module and no |games| or |colour| modules. You should
get the new package if you want to use the games worksheets.

This sheet needs some Python experience, but we've tried to make it as
easy to follow as possible.

\begin{note}
When we want to say something to more experienced programmers who are
reading this, we've enclosed it in a box like this.
\end{note}

\section{Beginners}

When you say |from livewires import *| at the start of a program (which
you usually should in the Beginners' worksheets), that makes all the
things in the |beginners| module available for your use.

\begin{note}
Saying |from livewires import *| causes all the things in |beginners| to
be imported into the local namespace. This is usually considered to be a
Bad Thing, but it is done in the Beginners' worksheets for the
sake of simplicity. We may change this at some point in the future.
We've chosen not to do it this way in the Games worksheets.
\end{note}

Some day, you might find yourself having to use Python without our
module. In that case, you'll need to know what's ours and what's in
standard Python. That's what this sheet is all about.

\subsection{Graphics}

Most of the |livewires| module deals with graphics. \emph{All} of the
graphics things discussed in the Beginners' worksheets depend on our
module. That doesn't mean you can't do graphics without our module; it's
just more difficult.

\begin{note}
Currently the graphical parts of the |beginners| module are implemented
as a wrapper round |Tkinter|. A canvas object holds all the other
objects on the screen.
\end{note}

\subsection{Input and output}

Just input, actually. The functions |read_number|, |read_string| and
|read_yesorno| come from |livewires|. You can get roughly the same
effect as |read_number| using the standard Python function |input|,
and exactly the same effect as |read_string| using the |raw_input|
function. If you need something like |read_yesorno| and can't use our
module, you'll need to write your own.

\subsection{Other things}

Our |random_between| function is exactly the same
as |random.random_int| in standard Python.

%-----------------------------------------------------------------------------

\section{Games}

The |games| module uses the ideas of classes and objects explained in
Sheet~O.  There are various classes in the module, which are explained
below.

In the Games worksheets, we've told you to use |from livewires import
games| when you want to access things in the |games| module. To use the
things in the module, you need to put |games| before the name. For
example, to refer to the |Screen| class in the |games| module, you'd
write |games.Screen|. 

In the |games| module, all co-ordinates are in pixels. (0,0) is the
top left corner. Co-ordinates increase downwards and to the right.
Times are in thousands of a second, that is, milliseconds. Colours are
specified in the way described in the |colour| module (see below).

\begin{note}
Most of these classes are designed to be extended by subclassing. To
avoid some of the syntactic pain that usually attends this kind of
design in Python, every class defines an extra method, which takes
exactly the same arguments as its |__init__| method. In a class called
|Foobar|, this method is called |init_foobar|.
\end{note}

The |games| module uses the Pygame library, which you can find at
\href{http://www.pygame.org/}{|http://www.pygame.org/|}. You need to
install the Pygame library before you can use the |games| module.

The games module is used by setting up objects which do what you want,
and then calling the |mainloop| method of the |Screen| class. The
|mainloop| method handles moving your objects around and updating the
screen. |mainloop| does not return until the player quits the game, so
all the behaviour of the game must go inside the objects you create
before calling |mainloop|.

Below, we list the classes in the |games| module, and their methods. For
each method, we give its arguments. An argument written as |name=value|
means the argument gets set to |value| if you don't specify it (so it's
not compulsory to specify it).

\subsection{Screen}

The |Screen| class provides a region in which objects can exist and
move.

The methods provided by the |Screen| class are listed below. We've
listed them as you would call them. If you make a subclass of the screen
class and want to redefine these methods, you'll need to add |self| as
the first parameter of the method.

\begin{itemize}

\item |init_screen (width=640, height=480)|

This method creates the window on the screen with the specified width,
height. You can only do this once in your program.

\item |is_pressed (key)|

This method returns 1 if the key is pressed, and 0 if it is not.

The key names are in the games module. The letter keys are |games.K_a|
to |games.K_z|. The space bar is |games.K_SPACE|. \key{Return} is
|games.K_RETURN|. If you import the games library at the |>>>| prompt
and say |dir (games)|, you can get a list of all the names in the games
library, which will include the keys. Or you could look at the Pygame
documentation at
\href{http://www.pygame.org/docs/ref/pygame_constants.html}{|http://www.pygame.org/docs/ref/pygame\_constants.html|},
as the key names are taken from the Pygame library.

{\it Note:} most keyboards can only tell you about 4 keys being pressed
a time, so once you're pressing more than 4 keys, |is_pressed| might
return 0 even for keys which are pressed.

\item |keypress (key)|

This method is another way of getting key presses. It is called
automatically whenever a key is pressed. In the standard |Screen| class,
it does nothing. If you override it in a subclass of |Screen| you can
get your subclass to handle key presses: the |key| parameter will be the
key which has just been pressed, using the same key names as the
|is_pressed| function above.

Whether you want to handle keys by writing your own |keypress| method or
using the |is_pressed| method depends on what you're doing: in all the
example sheets, other objects call the |Screen|'s |is_pressed| method to
find out whether a particular key is pressed, rather than the |Screen|
handling the key press itself.

\item |mouse_buttons ()|

In exactly the same way as |is_pressed|, this method returns the
pressed-ness of the mouse buttons.  It returns a {\it tuple} of three
numbers, being the state of the left, middle and right buttons
respectively.  The state is set to 1 if the button is pressed, and to
0 if it isn't.

{\it Note:} most Windows computers don't have a middle mouse button,
so you might not want to use it for anything important.

\item |mouse_position ()|

This method returns the current position of the pointer in the window
as an (x, y) pair.  You usually need this information if you're
handling a mouse click!

\item |mouse_down ((x, y), button)|

As with key presses, you can wait to be told about mouse clicks rather
than actively scanning for them.  This method is called automatically
whenever a mouse button is pressed.  In the standard |Screen| class,
it does nothing.  If you override it in a subclass of |Screen| you can
get your subclass to handle mouse clicks: the |x| and |y| parameters
get the position of the pointer on the screen, while |button| is the
number of the mouse button being pressed: 0 for left, 1 for middle and
2 for right.

Whether you want to handle mouse clicks by writing your own
|mouse_down| routine method or by using the |mouse_buttons| routine
instead depends on what you're doing.

\item |mouse_up ((x, y), button)|

This method is exactly like |mouse_down| except that it's called when
the mouse button is released.  For a lot of purposes you will want to
use this rather than |mouse_down|, unless speed is more important than
accuracy!

\item |set_background (background)|

This method sets the background of the screen. You need to give it an
image which you've loaded with |load_image|, not the filename of the
image. Note that the background image should not have transparency
set, or you'll get weird effects.

\item |set_background_colour (colour)|

Sets the background colour of the screen.  |colour| is a tuple of
three numbers between 0 and 255 indicating how much red, blue and
green there is respectively in the colour.  The |colours| module
provides useful constants for common colours.

\item |tick ()|

This method is called every timer tick (usually 1/50th of a
second). In the standard |Screen| class, it does nothing. If you make
a subclass of the |Screen| class, you can override this method to do
whatever you want to do every tick.

\item |handle_events ()|

This method enables you to write your own handler for the Pygame events,
rather than using the standard one which comes with the |Screen| class.
It is called every timer tick after all the objects have been updated.
You probably don't want to mess with this function unless you understand
Pygame's events module.

The standard handler calls |keypress| to deal with key down events,
|mouse_down| to deal with mouse down events, |mouse_up| to deal with
mouse up events, and |quit| to deal with quit events.

If you override this method, you {\em must} handle the quit event by
calling |self.quit ()|.

\item |quit ()|

Calling this method will stop the main loop from running and 
make the graphics window disappear. 

\item |clear ()|

Calling this method destroy all the |Object|s on the screen.

\item |mainloop (fps = 50)|

This method should be called once you've set up all the objects on the
screen. This method won't return until the screen's |quit| method is
called. It contains the loop which causes the objects on the screen to
be drawn and redrawn, so nothing will move on the screen before you
call it.

The |fps| parameter is the number of times the screen should be
updated every second. Whether or not your computer will actually
achieve this depend on how fast a computer you have.

\item |objects_overlapping (box)|

Returns a list of all the |Objects| on the Screen whose bounding boxes
overlap the box you give. A ``bounding box'' is a rectangle, with the
edges either horizontal or vertical, which completely encloses shape you
see on the screen.

The box should be given as a sequence of four elements, which are the
co-ordinates of the top left of the box, and its width and height,
like this: |(x0, y0, width, height)|.

\item |all_objects ()|

Returns a list of all |Object|s on the screen.

\end{itemize}

\subsection{Object}

The |Object| class represents a graphical object on the screen. You
shouldn't create |Object|s directly: the LiveWires Games module
provides subclasses of the |Object| class for you to use. Here
is a list of all the methods that the subclasses of |Object| have in
common:

\begin{itemize}

\item |destroy ()|

Removes the object from the |Screen|.

\begin{note}
This doesn't remove all references to the |Object| itself that you might
be keeping, so it doesn't guarantee that the |Object| will be removed
from memory. What it does do is remove any references the |games| module
is keeping to the |Object|.
\end{note}

\item |pos ()|

Returns position of the object's {\it reference point}. When we say
reference point, we mean some special point of the object: for a circle,
this is the centre. The description of each subclass of object will tell
you what its reference point is. 

The point is returned as a tuple: so you can say |(x, y) = my_object.pos
()|.

\item |xpos ()|

Returns the $x$-coordinate of the object's reference point.

\item |ypos ()|

Returns the $y$-coordinate of the object's reference point.

\item |bbox ()|
 
Returns a bounding box for the object.  A ``bounding box'' is a
rectangle, with the edges either horizontal or vertical, which
completely encloses shape you see on the screen.

The box is returned as a tuple of 4 numbers, which are the co-ordinates
of the top left of the box, and its width and height, like this: |(x0,
y0, width, height)|.

\item |move_to (x,y)| {\it or} |move_to ((x,y))|

Moves the object's reference point to (x,y), moving the object with it.

\item |move_by (dx,dy) {\it or} move_by ((dx,dy))|

Moves the object's reference point by |dx| in the $x$ direction and |dy|
in the $y$ direction, moving the object with it.

\item |rotate_to (angle)|

Sets the angle by which the object is rotated, in
anticlockwise degrees. An angle of 0 means the angle
at which the object was originally placed.

For some objects (circles, for example), this may actually do nothing,
except that the angle is remembered for returning from |angle ()|.

\item |rotate_by (angle)|

Adjusts the angle by which the object is rotated,
increasing it by "angle" degrees (anticlockwise).

For some objects (circles, for example), this may actually do nothing,
except that the angle is remembered for returning from |angle ()|.

\item |angle ()|

Return the object's current angle, in anticlockwise
degrees. The angle is always greater than or equal to zero, and less
than 360.

\item |overlaps (o)|

Returns true or false depending on whether this object
overlaps another object "o". 

\item |overlapping_objects ()|

Returns a list of the objects which are overlapping this object.

\item |filter_overlaps (object)|

You will not need to use this method unless you create your own
subclasses of |Object|.

This is a utility method which allows you to have better accuracy when
judging whether two objects have collided. Just checking whether the
bounding boxes which enclose them are overlapping will sometimes give
false collisions when the objects themselves do not overlap (assuming 
the objects aren't rectangles).

This method is called after having established that the bounding boxes
touch. 

Some subclasses of |Object| override it (e.g. the |Circle| class). You
can also override it in your own subclasses of |Object| to get better
collision detection.

This function should return 1 if the your object really overlaps the
other object and 0 otherwise. The standard |Object| class relies on the
bounding box check alone, so its |filter_overlaps| method just returns
1.

\item |raise_object (above=None)|

One way of viewing objects on a screen is with the objects being in
different layers, the front of the screen being the ``topmost''
layer.  Objects in ``higher'' layers then appear in front of (or
``above'') other objects.

This method brings the object towards the front of the screen, placing
it ``above'' the other object |above|.  If |above| is |None| or
missing, then the object is pulled to the topmost layer.

\item |lower_object (below=None)|

The reverse of |raise_object|, this puts the object behind |below|.
If |below| is missing or |None|, the object is put to the ``bottom''
of the screen, behind everything except the background.

\end{itemize}

\subsubsection{Colours}

|Polygon|, |Circle| and |Text| objects all have a defined colour, and
all have the following methods provided by the |ColourMixin| class:

\begin{itemize}

\item |set_colour (colour)|

This sets the main ``fill'' colour of the object.  |colour| is a tuple
of three numbers from 0 to 255, specifying how much red, green and
blue there is respectively in the colour.  The |colours| module
provides constants for many colours to avoid you having to experiment.

\item |get_colour ()|

This method returns the ``fill'' colour of the object.

\end{itemize}

\subsubsection{Outlines}

|Polygon| and |Circle| objects all have a defined outline colour, the
colour which will be used for the edge of the object.  They have the
following methods provided by the |OutlineMixin| class:

\begin{itemize}

\item |set_outline (colour)|

This sets the colour of the outline of an object.  |colour| is a tuple
of three red, green and blue values, as explained above, or |None|: if
|None| is specified, the ``fill'' colour of the object will be used
instead.

\item |get_outline ()|

This method returns the current outline colour of the object, or
|None| if no outline colour has been specified.

\end{itemize}

\subsection{Sprite}

This class represents an image you've loaded from a file, for example
the image of the asteroid in the Asteroids worksheet.  |Sprite| is a
subclass of |Object|, so it has all the methods from the |Object|
class as well as this:

\begin{itemize}
\item |init_sprite (screen, x, y, image, a=0, static=0)|

|screen| is the |Screen| which the image will be on. The centre of the
image will be at |(x,y)|, which is also the reference point.

|image| should be an image returned from the |load_image| function. See
below for more details on that function.

|a| is the angle of rotation you want the image to start at.

|static| is a flag that can be set to 1 if the sprite isn't intended
to move ever.  When there are a lot of static objects and not too much
else this can improve redrawing speed, but if there are only one or
two it can actually slow things down.

\end{itemize}

\subsection{Polygon}

This class represents a closed polygon on a |Screen|. When we say {\it
closed}, we mean that the lines making up the polygon completely enclose
a single area. A square is an example of a closed polygon. A square with
one side removed is not closed.

|Polygon| is a subclass of |Object|, so it inherits all the methods from
|Object|. It has these methods of its own:

\begin{itemize}

\item |init_polygon (screen, x, y, shape, colour, filled=1,
outline=None, static=0)|

|screen| is the |Screen| that the |Polygon| is on.

|x| and |y| are the co-ordinates of the reference point of the |Polygon|.

The |shape| is a list of pairs of co-ordinates |[(x0,y0), (x1,y1),
...]|, giving the $x$ and $y$ co-ordinates of the corners of the shape.
The co-ordinates are written relative to the reference point, that is,
when you're writing them, assume the reference point is at |(0,0)|. (See
the section on the ship in the ``Games: Space War'' sheet for an
example).

|colour| is the colour of the polygon.  This is a tuple of three
numbers from 0 to 255, specifying how much red, green and blue
respectively there is in the colour.  The |colours| module provides a
number of useful predefined colours.

If |filled| is 1, the polygon is filled in with its |colour|. If
|filled| is 0, only the outline of the polygon is drawn.

|outline| is the colour of the outline of the polygon, if it needs to
be different from the main |colour|.  If |outline| is |None|, then
|colour| is used instead.

|static| is a flag that can be set to 1 if the polygon isn't intended
to move ever.  It is mostly used for the |GridBox| class in the
|boards| module, which form the squares of a board.  When there are a
lot of static objects and not too much else this can improve redrawing
speed, but if there are only one or two it can actually slow things
down.

\item |set_shape (( (x0,y0), (x1,y1), ... ))| 

Change the shape of the object. The new shape is specified in exactly
the same way as the |shape| argument to |init_polygon|. The position
of the reference point on the screen stays the same.

\item |get_shape ()|

Return the current shape as a list of pairs of co-ordinates.

\end{itemize}

\subsection{Circle}

This class represents a circle on a |Screen|.  It is a subclass of
|Object|, so it inherits all the methods from |Object|. It has these
methods of its own:

\begin{itemize}

\item |init_circle (screen, x,y, radius, colour, filled=1,
  outline=None, static=0)|

|screen| is the |Screen| that the |Circle| is on.

The centre of the circle is at |(x,y)|. This is also the reference
point.

|radius| is the radius of the circle.

|colour| is the colour of the circle.

If |filled| is 1, the polygon is filled. If |filled| is 0, only the
outline of the circle is drawn.

|outline| is the colour of the edge of the circle, if it needs to be
different from the main |colour|.  If |outline| is |None|, then
|colour| is used instead.

|static| is a flag that can be set to 1 if the circle isn't intended
to move ever.  When there are a lot of static objects and not too much
else this can improve redrawing speed, but if there are only one or
two it can actually slow things down.

\item |set_radius (r)|

Sets the circle's radius to r.

\item |get_radius (r)|

Returns the circle's radius.

\end{itemize}

\subsection{Text}

This class represents some text on a |Screen|.  It is a subclass of
|Object|, so it inherits all the methods from |Object|. Notice that
the reference point is the centre of the text, which is not usually
what you expect.  |Text| has these methods of its own:

\begin{itemize}

\item |init_text (screen, x,y, text, size, colour, static=0)|

|screen| is the |Screen| that the |Text| is on.

|x| and |y| are the co-ordinates of the centre of the text.  The
reference point is |(x,y)|.

|text| is a string containing the text to be placed on the screen.

|size| is the height of the text.

|colour| is the colour of the text.

|static| is a flag that can be set to 1 if the text isn't intended
to move ever.  When there are a lot of static objects and not too much
else this can improve redrawing speed, but if there are only one or
two it can actually slow things down.

\item |set_text (text)|

Sets the current text.

\item |get_text ()|

Returns the current text as a string.

\end{itemize}

\subsection{Timer}

The |Timer| class is a class you can add to something which is also a
subclass of |Object|, to make an object that performs actions at regular
intervals. A class which is intended to be used with another class is
called a ``mix-in''. For instance, if you wanted to make a new class of
your own which was a |Circle| and also a |Timer|, you would define the
class by saying |class MyClass (games.Circle, games.Timer):|

\begin{itemize}
\item |init_timer (interval, running=1)|

|interval| is how often the |tick| method is called, measured in timer
ticks. How long a tick is depends on the |fps| argument you give to the
|Screen|'s |mainloop| method. Setting |fps| to 50 means a tick is 1/50
of a second.

|running| controls whether the timer starts immediately on
initialisation, or whether you need to explicitly |start| it later.

You must call this method \emph{after} you have called the |init_| method for
the |Object| subclass you are using. 

\item |stop ()|

Stop the timer running. It continues to exist, but doesn't count any
more.

\item |start ()|

Starts the timer again. A full interval will elapse before it ticks.

\item |get_interval ()|

Gets the current interval.

\item |set_interval (interval)|

Sets the current interval.

\item |tick ()|

This method must be {\it supplied} by you, by subclassing the |Timer|
class.  If it is not present, the module will generate an error and
quit.

\end{itemize}

\subsection{Mover}

The |Mover| class is a subclass of the |Timer| class. A |Mover| is
something which moves itself around the screen at regular intervals.

Like the |Timer| class itself, the |Mover| class is intended to be used
as a mix-in class with a subclass of |Object|. For example, to create
your own class representing a moving polygon, you would say |class Ship
(games.Polygon, games.Mover)|

\begin{itemize}
\item |init_mover (dx, dy, da=0)|

You must call this method {\it after} you have called the |init_| method for
the |Object| subclass you are using.

Every tick (see |Timer|, above), your object will |move_by| |dx| pixels in the
$x$ direction and |dy| pixels in the $y$ direction. If |da| is given and
is non-zero the object will also be rotated by |da| degrees. The
object's |moved ()| method will then be called.

Note that your object must {\it have} a |moved| method, so you must
provide one in your subclass.

\item |set_velocity ([dx,dy])| {\it or} |velocity ((dx,dy))|

This method sets |dx,dy|.

\item |get_velocity ()|

This method returns |(dx,dy)|

\item |set_angular_speed (da)| 

This method sets |da|. 

\item |get_angular_speed ()|

This method returns |da|.

\item |moved ()|

You must supply the |moved| method by subclassing. The |moved| method is
called on each tick after the object has been moved by |(dx,dy)| and
rotated by |da|.

\emph{Note:} if all of |dx|, |dy|, and |da| are zero, this method is still
called each tick.

\end{itemize}

\subsection{Message}

The |Message| class is a subclass of the |Text| class and the |Timer|
class. A |Message| is a |Text| object that will go away after a
specified period.

\begin{itemize}

\item |init_message (screen, x, y, text, size, colour, lifetime,
after_death=None)| 

Except for the last two arguments, this method works identically to the
|init_text| method (see the |Text| class we mentioned previously).

|lifetime| is the lifetime of the message in ticks, that is, how
long it will last before it goes away.

When the message goes away, the |after_death| argument will be called if
it was supplied. If it is supplied, the |after_death| argument must be
something which is callable (a function, for example). 

\end{itemize}

\subsection{Animation}

The |Animation| class is a subclass of the |Sprite| and |Timer| classes.
It allows you to produce animations, that is, pictures on the screen
which change every tick.

You construct an |Animation| by saying |Animation (screen, x, y,
nonrepeating_images, repeating_images, n_repeats, repeat_interval)|, where

|screen| is the screen the animation is on.

|(x,y)| are the co-ordinates of the centre of the animation.

|nonrepeating_images| is a list of images from |load_image| or a list of
filenames.

|repeating_images| is a list of images from |load_image| or a list of
filenames.

|n_repeats| is the number of times the images in the |repeating_images|
list will be shown. Set it to -1 to keep repeating forever.

|repeat_interval| is the number of ticks (see the |Timer| class) between
one image and the next.

If the |nonrepeating_images| list is $[a,b,c,d]$ and the
|repeating_images| list is $[x,y,z]$ then the sequence of images shown
will be $a,b,c,d, x,y,z, x,y,z, x,y,z, ...$.

The |Animation| class isn't currently intended to be subclassed: this
may change in future.

\subsection{Useful functions in the games module}

\begin{itemize}

\item |load_image (file, transparent=1)|

Loads an image from a file and returns an object which can be passed to
the |games| modules' classes as the image for a |Sprite| or
|Background|.

|file| is a string containing the name of the file to load.

|transparent| indicates whether the image should be transparent. If it
is not zero, the image is transparent. The transparent colour is taken
from the top left corner of the image. You should not make transparent
images which you intend to use as the background for the screen.

\begin{note}
The object returned by |load_image| is a |pygame.Surface| object.
\end{note}

\item |load_sound (file)| 

Loads a sound object from a WAV file and returns an object which you can
use to play the sound.

You can use it in the following sort of way:

\begin{program}
mysound = games.load_sound ('explosion.wav')
mysound.play ()
\end{program}

The object returned by |load_sound| is a |pygame.mixer.Sound| object.

For more information on the methods of sound objects see the documentation
for Pygame at\\
|\href{http://www.pygame.org/docs/ref/Sound.html}{http://www.pygame.org/docs/ref/Sound.html}|.

\item |load_animation (nonrepeating_files, repeating_files=[], transparent=1)|

Loads a sequence of images from files like |load_image|, returning a
pair of lists: first all the |nonrepeating| images, then all the
|repeating| images (if any).  These lists of images are suitable for
making an |Animation|.

\item |scale_image (image, x_scale)|
{\it or}
|scale_image (image, x_scale, y_scale)|

This function scales an image object from |load_image| by the amount you
specify.

If you give it one scale factor, the image is scaled by the same amount
in the $x$ and $y$ directions (giving a factor of 2.0 would double the
size of the image, for example).

If you give it two scale factors, you can scale the image by different
amounts in the $x$ and $y$ directions.

This function returns the scaled image object. It does not modify the
original image you gave it.
\end{itemize}

%-----------------------------------------------------------------------------

\section {Boards}

The |boards| module is an extension of the |games| module, providing
subclasses and useful functions particularly aimed at writing board
games.  The definition of ``board games'' is quite wide -- the
|boards| module can equally well be used for Snake as for Draughts!

In the Games worksheets, we've told you to use |from livewires import
boards| when you want to access things in the |boards| module. You
will usually want to access things in the |games| module also.  To use
the things in the module, you need to put |boards| before the
name. For example, to refer to the |SingleBoard| class in the |boards|
module, you'd write |boards.SingleBoard|.

\begin{note}
These classes follow the same conventions as the |games| module.
Coordinates are given relative to the top left hand corner of the
window, for example, and any class called |FooBar| has an extra method
called |init_foobar|.
\end{note}

Most of the comments about the games module apply to the boards module
too.  In particular, the |SingleBoard| class is a subclass of
|Screen|, so has a |mainloop| method that handles moving everything
around the screen.

Below, we list the classes in the |boards| module, and their methods. For
each method, we give its arguments. An argument written as |name=value|
means the argument gets set to |value| if you don't specify it (so it's
not compulsory to specify it).

\subsection{GameCell}

|GameCell| is a subclass of |Polygon| from the |games| module.  It is
intended to be subclassed itself to fit the exact needs of a
particular game, since it provides only fairly general methods and
properties.  Note, however, that a |GameCell| is a {\em static}
Polygon; you will often need to mix the |Container| class below into
your subclass to get it to redraw properly.  |GameCell|'s shape is
always a square.

Primarily, the |GameCell| class provides a number of useful
properties, variables that belong to each |GameCell| separately.
These properties are:

\begin{itemize}

\item |grid_x|

The column number of the cell in the grid, counting from zero as the
leftmost column.

\item |grid_y|

The row number of the cell in the grid, counting from zero as the top
row.

\item |neighbours|

A list of all the |GameCell|s which are next to this |GameCell|.
These lists are created by a method in the |GameBoard| class, which
allows some control over exactly what ``neighbour'' should be taken to
mean.

\item |direction|

An array containing the |GameCell| which can be found in each
direction from this |GameCell|.  This array is created by a method in
the |GameBoard| class, which allows some control over exactly what
``direction'' should mean.

\end{itemize}

The library supplies a set of constants for using the |direction|
array in a meaningful way.  These are:

\begin{itemize}
\item |boards.LEFT|
\item |boards.UP_LEFT|
\item |boards.UP|
\item |boards.UP_RIGHT|
\item |boards.RIGHT|
\item |boards.DOWN_RIGHT|
\item |boards.DOWN|
\item |boards.DOWN_LEFT|
\end{itemize}

You should not rely on these constants being particular values or in a
particular order.  Instead, the module provides a number of useful
functions to turn one direction into another, described later on.

In addition to the methods inherited from |Polygon|, the following
methods can be used on a |GameCell|:

\begin{itemize}

\item |init_gamecell (board, i, j, line_colour=colour.black,|\\
|\verb+               +fill_colour=colour.light_grey)|

|board| is the |GameBoard| (see below) that the |GameCell| is on.

|(i, j)| is the grid coordinates of the cell.  In other words, |i| is
the column number of the cell within the grid, while |j| is the row
number.  Cell |(0, 0)| is the top left-hand corner of the grid.

|line_colour| is the colour of the line drawn between squares in the
grid (i.e. the outline colour of the |Polygon|).

|fill_colour| is the colour of the square itself.

\item |add_neighbour(neighbour)|

You probably won't need to call this method yourself, but if you do
this is how.  This method is called by the |GameBoard| class to create
the |neighbours| list for a |GameCell|.  |neighbour| is a |GameCell|
which is next to (in some sense) this cell.

{\it Note:} this only sets up a one way relationship.  In other words,
the instruction |fred.add_neighbour(bill)| tells |fred| that |bill| is
his neighbour, but doesn't tell |bill| anything at all about |fred|.

\item |add_direction(dir, cell)|

Again, you probably won't need to call this method yourself since it
is called by the |GameBoard| class.  It builds up the |direction|
array for this |GameCell|.

|dir| is the direction under consideration.

|cell| is the |GameCell| to be found in direction |dir|.

{\it Note:} again, this method only establishes a connection one way.
The instruction |lobby.add_direction(LEFT, main_hall)| tells the
|lobby| that the |main_hall| is to its left, but doesn't let on to
the |main_hall| that the |lobby| is near it at all.

\end{itemize}

\subsection{GameBoard}

The |GameBoard| class is where you keep the |GameCell|s on your
board.  You will almost always want to subclass |GameBoard| (or more
likely the |SingleBoard| class described below) to get exactly what
you want for your game.

Each |GameBoard| contains a rectangular grid of |GameCell|s.  Not all
cells in the grid have to actually contain |GameCell|s, but all the
cells are assumed to be square whether or not they are present.

The |GameBoard| class provides two useful properties (variables):

\begin{itemize}

\item |grid| is a list of lists (see Sheet~A if you're not sure what
that means) which contains every |GameCell| in the table.  The cell at
|grid[i][j]| is the cell which will appear in column |i|, row |j| on
the screen.

\item |cursor| is an optional feature of the library.  If it is
switched on using |enable_cursor| (see below), |cursor| contains the
|GameCell| which is currently highlighted.  All the boardgame
worksheets use |cursor| to control the game from the keyboard, moving
the ``cursor'' around the board to select a particular cell that some
other action should happen in.

\end{itemize}

|GameBoard| provides the following methods:

\begin{itemize}

\item |init_gameboard (screen, (x, y), n_cols, n_rows, box_size,|\\
|\verb+                +line_colour=colour.black, fill_colour=colour.light_grey,|\\
|\verb+                +cursor_colour=colour.red)|

|screen| is the |Screen| on which this board lives.

|(x,y)| are the coordinates of the top left-hand corner of the board;
under most circumstances these will be the coordinates of the top
left-hand corner of the |GameCell| in the top left-hand corner of the
grid.

|n_cols| and |n_rows| are the number of columns and rows of
|GameCell|s in the grid respectively.

|box_size| is the height and width of each square in the grid.

|line_colour| is the colour of the lines between or around each
|GameCell| (in other words, the |outline| colour of the |Polygon|).

|fill_colour| is the colour of each |GameCell|.

|cursor_colour| is the colour of the ``cursor'', the colour which is
used instead of the line colour to highlight a particular |GameCell|
if this feature of the class is used.

\item |new_gamecell (i, j)|

You will probably want to override this method in any subclass that
you create.  This method is called automatically to create a new cell
to fit into column |i|, row |j| of the grid.  The version supplied
with |GameBoard| itself creates a |GameCell|, which is almost
certainly not what you want.  Instead you should write a version which
returns one of the subclasses of |GameCell| that you have made.

\item |create_neighbours(orthogonal_only=0)|

This method must not be called before |init_gameboard| has been
called.  It scans through the |grid|, informing each |GameCell| of
which other |GameCell|s are its neighbours.

|orthogonal_only| is a flag controlling whether or not to include
cells that are only diagonally adjacent to one another.  If it is set
to 0, cells on the diagonal are included; if it is 1, then only the
cells directly up, right, down or left ({\it orthogonally adjacent})
of each cell are considered to be neighbours.

\item |create_direction(orthogonal_only=0, wrap=0)|

This method must not be called before |init_gameboard| has been
called.  It scans through the |grid|, informing each |GameCell| of
which other |GameCell|s lie in which directions from it.

|orthogonal_only| is a flag controlling whether or not the diagonal
directions are filled in.  If it is 0 then the diagonals are
considered, otherwise they are not.  This is not usually a major
concern, and may be removed in future versions of the library.

|wrap| is a flag which controls whether or not to ``wrap'' the board's
sense of direction.  If |wrap| is 0, then there is considered to be
nothing to the left of the leftmost column of cells, nothing above the
top row of cells, and so on.  If |wrap| is 1, then the rightmost
column of cells are considered to be to the left of the leftmost
column of cells, and the bottom row of cells are considered to be
above the top row of cells, and vice versa.  This allows a piece to
move off the top of the board onto the bottom, for example.

\item |keypress (key)|

This method should be called whenever a key is pressed.  This will
happen automatically in any class that subclasses both |GameBoard| and
|Screen|, such as |SingleBoard|.

The method checks to see if it understands the key code in |key| as an
instruction to move the cursor.  If it doesn't, it calls
|handle_keypress|, allowing the game to do any special processing that
it might need to do.

Key codes that |keypress| understands are held in the property
|key_movements|.  This is a dictionary containing key codes matched to
|(di, dj)| pairs, where |di| is the required change in the column
number and |dj| is the change in row number for the cursor.  The
|keypress| method does not allow these changes to take the cursor off
the board, nor does it attempt to ``wrap'' the cursor by reintroducing
it on the right if it tries to fall off the left of the grid.  The key
codes and movements set up by default are:

\begin{itemize}
\item |K_UP|: (0, -1)
\item |K_DOWN|: (0, 1)
\item |K_LEFT|: (-1, 0)
\item |K_RIGHT|: (1, 0)
\end{itemize}

\item |handle_keypress(key)|

This method is called automatically whenever the |boards| library
receives a key code that it does not know how to handle.  Except as
noted above, it is used in exactly the same way as
|Screen.handle_keypress| in the |games| module.  You should override
this method in your subclass if you want to handle keyboard input
yourself.

\item |enable_cursor(i, j)|

This method switches on the cursor, placing it on the cell in column
|i|, row |j|.  If the cursor already exists, it is simply moved to
|(i,j)|.  If |(i, j)| is not a valid cell on the grid, an error will
occur.

\item |disable_cursor()|

This method switches off the cursor, setting |cursor| to |None| and
removing the red outline from the screen.  It is safe to call this
method if the cursor is already disabled --- in that case, nothing
will happen.

\item |move_cursor(i, j)|

This method moves the cursor to the cell at column |i|, row |j| on the
grid.  The cursor must already be enabled, and |(i, j)| must be a
valid cell on the grid, otherwise an error will occur.

\item |cursor_moved()|

This method is called automatically whenever the cursor is enabled,
disabled (provided it was enabled at the time) or moved either
explicitly in the program or by pressing keys.  By default, it does
nothing --- you should override this method in your subclass if you
want to be notified when the state of the cursor has changed.

This method may change in the future, since it's not enormously useful
at present.

\item |on_board(i, j)|

This method returns |1| if grid position |(i, j)| is really on the
board, and |0| otherwise.  By default, |(i, j)| is considered to be on
the board if both |i| and |j| are zero or more, |i| is less than the
maximum number of columns passed to |init_gameboard| and |j| is less
than the maximum number of rows.

|on_board| is used when creating the grid, and when setting up the
|direction| and |neighbours| properties.  You can override it in your
subclass to produce a board that has holes in it (a Ludo board, for
instance) automatically.

\item |map_grid(fn)|

|fn| is a callable object (a function, usually) which is expected to
take an entry from |grid| as a parameter.  The |map_grid| method then
calls |fn| on every cell on the board.  There is an example of how to
use it in the ``Games: Minesweeper'' sheet: it is most often handy for
tidying up at the end of a game.

\item |cell_to_coords(i, j)|

This method returns the pixel coordinates (the |(x,y)| used by the
|games| module) of the top left-hand corner of the cell in column |i|,
row |j|.

\item |coords_to_cell(x, y)|

This method returns the grid coordinates (the |(i,j)| used by the rest
of the |boards| module) of the cell inside which the point |(x,y)| in
pixel coordinates lies.  It is most often used for working out which
|GameCell| the pointer was over when a mouse button was pressed or
released.  {\em Note:} the result is not necessarily a legal position
for a cell, so check it with |on_board| before using it!

\end{itemize}

\subsection{SingleBoard}

The |SingleBoard| class is a straightforward subclass of |GameBoard|
and |Screen|.  It provides a single grid of |GameCell|s in a window,
which is the basis of most games.  This is the class that you will
probably want to subclass for your game, unless for some reason you
want to have two or more boards in your window.

|SingleBoard| provides one method in addition to those it inherits
from |Screen| and |GameBoard|:

\begin{itemize}

\item |init_singleboard(margins, n_cols, n_rows, box_size,
line_colour=colour.black,|\\
|\verb+                 +fill_colour=colour.light_grey,
cursor_colour=colour.red)|

|margins| is a tuple giving the amount of space to leave on each side
between the edge of the window and the edge or the grid.  It comes in
two alternatives: either |(x,y)|, where |x| is the amount of space at
the left and right and |y| the space above and below, or |(l,t,r,b)|,
where |l| is the left margin, |r| the right margin, |t| the space at
the top and |b| the space at the bottom.

|n_cols| and |n_rows| are the number of columns and rows respectively
in the grid.

|box_size| is the height and width of the grid squares.

|line_colour| is the colour of the lines between or around each
|GameCell| (in other words, the |outline| colour of the |Polygon|).

|fill_colour| is the colour of each |GameCell|.

|cursor_colour| is the colour of the ``cursor'', the colour which is
used instead of the line colour to highlight a particular |GameCell|
if this feature of the class is used.

\end{itemize}

\subsection{Container}

The |Container| class is a utility class that can be mixed in when you
subclass any |Object| subclass.  Its purpose is to keep together, in
the right order, any |Object|s that are supposed to stay together.
For example, if a counter is represented on the screen by a white
|Circle| with a small red |Circle| in the middle of it, the
|Container| class allows us to move the white |Circle| and have the
red |Circle| tag along automatically.

|Container| overrides a number of the normal methods of an |Object|.
As such, it must be the first class that you list in the brackets when
making a subclass.  For example:

\begin{program}
class MyCounter(boards.Container, games.Circle):
\end{program}

will work while

\begin{program}
class MyCounter(games.Circle, boards.Container):
\end{program}

will quietly and frustratingly fail to do what you want.

As a side-effect of |GameCell|s being declared to be |static| (q.v.),
you will usually want to make your game's subclass of |GameCell| a
subclass of |Container| as well.  In fact the only board game
worksheet for this isn't true at present is ``Snake'', which isn't
strictly speaking a board game at all!

|Container| provides one method for general use:

\begin{itemize}

\item |init_container(contents)|

|contents| is slightly unusual: it is a list of the names of the
objects which are to be contained.  For example, suppose we have the
white |Circle| containing a red |Circle| that we mentioned above, and
we set it up like this:

\begin{program}
self.red_blob = games.Circle(blah, blah, colour.red)
\end{program}

Then we would have to call |init_container(['red_blob'])| to
initialise it.  Suppose our counter had to have a number on it as
well:

\begin{program}
self.digit = games.Text(blah, blah, "1", blah)
\end{program}

Then we would need to call |init_container(['red_blob', 'digit'])|.
This would make sure that both the |red_blob| and the |digit| stayed
on top of our counter.  Furthermore, |digit| would be on top of
|red_blob|!

Note that the ``contents'' don't have to overlap the |Container| or
each other, but they usually do.  There is also nothing to stop the
contents being a |Container| themselves; for an example of that, see
the worksheet ``Games: Chain Reaction''.

\begin{note}
Note that the |init_container| method must be called before the init
routine for any other subclass (|init_circle| in the example above),
otherwise the library will produce a cryptic |KeyError| and quit.
\end{note}

As a side-effect, |init_container| will create |self.red_blob| and
|self.digit|, setting them to |None|.  You should therefore not create
your objects until after you have initialised your container!

\end{itemize}

\subsection{Useful functions in the boards module}

The module supplies a number of useful functions specifically for
playing with directions.  You should always use these functions rather
than making any assumptions about what numbers mean what directions.
That way, if the library ever changes you won't have to retype
everything!

\begin {itemize}

\item  |turn_45_clockwise(direction)|

This function returns the direction 45 degrees clockwise of
|direction|.  In other words, if you call it with |UP|, it will reply
with |UP_RIGHT|.

The remaining functions all follow the same naming scheme, apart from
the last one.

\item |turn_45_anticlockwise(direction)|
\item |turn_90_clockwise(direction)|
\item |turn_90_anticlockwise(direction)|
\item |turn_180(direction)|

\item |random_direction(orthogonal_only=0)|

To avoid the need for you to write lots of tedious |if|s or cunning
look-up code, this function returns a random direction.  If
|orthogonal_only| is 0, all 8 directions are equally likely to be
selected; otherwise only the {\it orthogonal} directions (|UP|,
|DOWN|, |LEFT| and |RIGHT|) will be chosen from.

\end{itemize}

%-----------------------------------------------------------------------------

\section {Colour}

The |colour| module provides some pre-defined colours for use with the
|games| and |boards| modules: whenever the modules wants a colour, you
can give them an object from this module.

The colours are:

\begin{itemize}
\item |red|
\item |green|
\item |blue|
\item |black|
\item |white|
\item |dark_red|
\item |dark_green|
\item |dark_blue|
\item |dark_grey|
\item |grey|
\item |light_grey |
\item |yellow|
\item |brown|
\item |pink|
\item |purple|
\end{itemize}

To use the module, say |from livewires import colour|. You can now refer
to the colours as |colour.red|, |colour.green| and so on.

If you want to make your own colours, you'll need to know that 
the colours are a tuples of 3 numbers giving the amount of red, green
and blue in the colour. The numbers are in the range 0 to 255. For
example the |colour.red| is |(255,0,0)|, giving all red, no blue and no
green.

The colours from the |colour| module are not intended for use with the 
|beginners| module.

\section{Who we are}

This section contains some background about us and LiveWires.

\subsection{Contributors}

Gareth McCaughan and Richard Crook both specified and wrote the
LiveWires package, with the assistance of Neil Turton and Matthew
Newton.  Paul Wright ported the games library to use Pygame.  Rhodri
James added the boards library.  Gareth wrote most of the Beginners'
worksheets. Rhodri, Neil and Paul wrote various combinations of
Beginners' and Games sheets.  Mark White wrote the |wsheet| \LaTeX{}
class which enabled us to produce the Postscript and PDF versions of
the sheets.

The rest of the team kept us sane on the LiveWires holiday itself. On
the LiveWires 2001 holiday, the rest of the computing team was Rob
Pearce and Colin Bell.

The maintainers of the course can be reached at \href{mailto:python@livewires.org.uk}{|python@livewires.org.uk|}.

\subsection{LiveWires}

LiveWires is a Scripture Union holiday for 12 to 15 year olds, which
takes place in the UK every summer. The young people on the holiday have
the chance to take part in a variety of computing, electronics and
multimedia activities. The LiveWires Python Course was written by to
help us to teach Python to the young people. We're making it available
to everyone else as a way of giving something back to the Python
community.

The LiveWires web site is at
\href{http://www.livewires.org.uk/}{|http://www.livewires.org.uk/|}

\subsection{Scripture Union}

Scripture Union is an organisation whose aim is to make Jesus known to
children, young people and families. SU staff and volunteers work in
more than 130 countries; in the UK its work includes schools work,
missions, family ministry, helping Christians to read the Bible and
supporting the church through training and resources. Scripture Union
holidays have been happening for more than 100 years. 

For more information on SU, see \href{http://www.scriptureunion.org.uk/}{|http://www.scriptureunion.org.uk/|}

\end{document}
